---
title: "Clade waves — concepts you will use"
subtitle: "Time-calibrated trees, Nextstrain metadata, and tip color in `ggtree`"
author: "Instructor notes"
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 9
    fig-height: 6
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Render from the project root with:
#   quarto render notes/concepts.qmd
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r packages}
library(ape)
library(treeio)
library(ggtree)
library(ggplot2)
library(dplyr)
library(readr)
library(lubridate)
```

# Why this document exists

This is a small, self-contained gallery of the techniques you need for the **clade-waves** starter project. Every concept is illustrated with a plot generated from the actual project data (`data/ncov_subsample.nwk` + `data/ncov_metadata.tsv`). Read it once before you start so you know what a finished figure looks like; come back to it when you forget which `ggtree` argument controls the time axis.

This starter focuses on a single question: *did SARS-CoV-2 clades replace each other through time?* You only need three tools — a phylogeny, a metadata table, and `geom_tippoint` — but using them correctly is the whole game.

# 1. A phylogenetic tree is a 2-D coordinate system

This is the single most important idea.

A drawn tree has two axes:

- The **y-axis** indexes tips. Internal y-coordinates are the average of the children's y-coordinates. The order is determined by `ladderize()` and is otherwise arbitrary.
- The **x-axis** carries either **substitutions per site** (the raw output of a tree-builder) or **calendar time** (the output of a clock-calibration step). For SARS-CoV-2 builds, it is calendar time.

Once you accept this, *anything else you draw is just another layer on the same coordinate system*: tip points at `(x_tip, y_tip)`, abundance bars elsewhere, … The whole figure is one big `ggplot`.

# 2. Loading the data

```{r load_data}
tree <- read.tree("data/ncov_subsample.nwk")
meta <- read_tsv("data/ncov_metadata.tsv", show_col_types = FALSE)

cat("Tips in the tree:", length(tree$tip.label), "\n")
cat("Date range:",
    as.character(min(meta$date)), "→",
    as.character(max(meta$date)), "\n")
cat("Distinct Nextstrain clades:", length(unique(meta$Nextstrain_clade)), "\n")
```

# 3. The join-key invariant — run BEFORE you join

`ggtree` binds metadata to a tree by matching `tree$tip.label` to a `strain` column. If even one strain name is mangled (an extra space, a `/` replaced by `_`), the match silently misaligns and your figure is wrong without warning.

Always run this two-line check first:

```{r join_check}
# These two should be: the first empty, the second potentially non-empty (fine).
length(setdiff(tree$tip.label, meta$strain))
length(setdiff(meta$strain, tree$tip.label))
```

If the first line is anything other than `0`, **stop and fix the strain naming on both sides** with the same regex before doing anything else.

# 4. Branch length units — verify before plotting on a time axis

`ggtree(td, mrsd = ...)` only works if branches are in **time** (years). If they are in substitutions/site, the time axis labels will be silently wrong.

```{r units_check}
max(node.depth.edgelength(tree))
```

Interpretation:

- `≈ 3–6` → branches are in **years** (time-calibrated). You can use `mrsd` directly.
- `≈ 0.001–0.01` → branches are in **substitutions/site**. You would need to time-calibrate first (not needed here — `data/ncov_subsample.nwk` is in years).

# 5. Nextstrain metadata anatomy — only the columns you need

The metadata file has 21 columns; for this starter you need just five:

```{r meta_anatomy}
meta %>%
  select(strain, date, decimal_date, Nextstrain_clade, pango_lineage) %>%
  slice_sample(n = 6)
```

| Column            | What it means                                                              |
|-------------------|----------------------------------------------------------------------------|
| `strain`          | Join key. Must match `tree$tip.label` exactly.                             |
| `date`            | Sampling date as `YYYY-MM-DD`.                                             |
| `decimal_date`    | Same date as a numeric (e.g. `2021.456`). Used as the `mrsd` anchor.       |
| `Nextstrain_clade`| Coarse-grained lineage label (`19A`, `20I (Alpha)`, `21K (Omicron)`, …).   |
| `pango_lineage`   | Finer-grained Pango notation (`B.1.1.7`, `BA.1`, `JN.1`, …).               |

A Nextstrain clade is the standard unit of "variant" reporting. The clade column is what you bind to tip color.

```{r clade_counts}
meta %>%
  count(Nextstrain_clade, sort = TRUE) %>%
  head(10)
```

# 6. The base time-axis tree

Every figure in this project is built on top of this single object.

```{r base_tree}
meta <- meta %>%
  mutate(decimal_date = decimal_date(ymd(date)))

td <- full_join(as_tibble(tree), meta, by = c("label" = "strain")) %>%
  as.treedata()

mrsd_value <- max(meta$decimal_date, na.rm = TRUE)

p_base <- ggtree(td,
                 mrsd = mrsd_value,
                 as.Date = TRUE,
                 size = 0.25,
                 color = "grey40") +
  theme_tree2() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Foundation: time-calibrated tree, 400 SARS-CoV-2 tips, 2020–2026")

p_base
```

What the arguments do:

- `mrsd = mrsd_value` — the **most recent sampling date**. `ggtree` subtracts branch lengths from this anchor to place every internal node on the calendar.
- `as.Date = TRUE` — formats x-axis values as `Date` objects so `scale_x_date` works.
- `size`, `color` — branch line width and color. Subtle defaults make the tip points stand out later.

# 7. Tip aesthetics — the answer figure

The whole question — "did clades replace each other?" — is answered by adding `geom_tippoint(aes(color = Nextstrain_clade))` to the base tree:

```{r tip_color, fig.height=8}
p_clade <- p_base +
  geom_tippoint(aes(color = Nextstrain_clade), size = 1.4, alpha = 0.85) +
  scale_color_viridis_d(option = "turbo", name = "Clade") +
  guides(color = guide_legend(override.aes = list(size = 3),
                              ncol = 2)) +
  ggtitle("Tips colored by Nextstrain clade — the wave pattern")

p_clade
```

Read the figure horizontally: tips at the same calendar date are vertically stacked, so a *band of color* at one date means many tips of the same clade at that date. The transitions between bands are the **clade waves** you are being asked to interpret.

Things to notice in this dataset:

- `19A` and `19B` (the original Wuhan-era lineages) sit at the very left of the plot, near 2020.
- `20I (Alpha)`, `21A (Delta)`, `21J (Delta)` form a left-to-right succession through 2020–2021.
- The 2022–2023 column is dominated by `22B (BA.5)` and `21L (BA.2)`.
- The 2024–2026 right edge is `24A (JN.1)`, `25B (NB.1.8.1)`, `25C (XFG)` — Omicron descendants that have moved past the original BA.x.

You should be able to point at *specific transitions on this figure* in your written interpretation.

# 8. Same data, no tree — clade abundance over time

Sometimes the cleanest way to communicate the wave pattern is *without* the tree. A stacked-area plot of clade fraction over time is a frequent companion figure:

```{r clade_area, fig.height=5}
clade_by_month <- meta %>%
  mutate(month = floor_date(ymd(date), "month")) %>%
  count(month, Nextstrain_clade) %>%
  group_by(month) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup()

ggplot(clade_by_month, aes(x = month, y = fraction, fill = Nextstrain_clade)) +
  geom_area(position = "stack") +
  scale_fill_viridis_d(option = "turbo", name = "Clade",
                       guide = guide_legend(ncol = 2)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = "Fraction of monthly sample",
       title = "Clade composition by month — the same waves, no tree") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Both figures answer the same question. The tree version preserves *evolutionary* relationships (a viewer can see that BA.5 and JN.1 share an ancestor); the area plot strips that information away to focus on *temporal succession*. Pick whichever makes your interpretation easier to write — or include both.

# 9. Pitfalls quick-reference

```{r pitfalls, echo=FALSE, results='asis'}
tab <- tribble(
  ~Symptom, ~Cause, ~Fix,
  "X-axis labelled in years but events are pre-1900",
    "Branches in subs/site, not time",
    "`max(node.depth.edgelength(tree))` should be a few years; otherwise time-calibrate",
  "Half the tips are NA after joining metadata",
    "Strain-name encoding mismatch (`/`, `|`, whitespace)",
    "Run `setdiff(tree$tip.label, meta$strain)` before joining",
  "Tip color legend is illegible (15+ entries crammed into one column)",
    "Default `guide_legend()` doesn't wrap",
    "`guides(color = guide_legend(ncol = 2))`",
  "Clade colors are visually too similar",
    "`Set1` / `Dark2` palettes max out around 8 colors",
    "`scale_color_viridis_d(option = \"turbo\")` handles 15+ categories"
)
knitr::kable(tab)
```

# 10. Where to go next

When this project is done, the natural next steps live in sibling directories:

- **`SARS-CoV_mutation_heatmap/`** — same tree, different question: *did key spike mutations evolve convergently?* You add `gheatmap` to the figure you built here.
- **`SARS-CoV_analysis_project/`** — the full version, where you layer all six visual encodings (tip color, branch color, heatmap, facet strips, MSA, geographic projection) on the same tree.

Each of these reuses the foundation you build in this starter.

# Session info

```{r sessioninfo, echo=FALSE}
sessionInfo()
```
