---
title: "Introduction to Base Graphics in R"
format:
  revealjs:
    embed-resources: true
editor: visual
---
<style>
.reveal {
  font-size: 160%;
}
</style>

# Base Graphics

- `plot()` — scatter plots, line plots, and more
- `hist()` — histograms
- `boxplot()` — box plots
- `barplot()` — bar charts
- `stripchart()` — one-dimensional scatter plots
- `par()` — graphical parameters and multi-plot layouts

## Scatter Plot

```{r}
#| echo: true
#| include: true
set.seed(123)
x <- rnorm(50) + 1:50
y <- rnorm(50) + 1:50

plot(x, y,
     main = "Scatter Plot",
     xlab = "X-axis",
     ylab = "Y-axis",
     pch = 19,
     col = "blue")
```

## Line Plot

```{r}
#| echo: true
#| include: true
x_line <- 0:10
y_line <- x_line^2

plot(x_line, y_line,
     type = "l",
     main = "Line Plot",
     xlab = "X-axis",
     ylab = "Y-axis",
     col = "red")
```

## Plot types: `type` argument

```{r}
#| echo: true
#| include: true
par(mfrow = c(2, 2))
plot(x_line, y_line, type = "l", main = "type = 'l'")
plot(x_line, y_line, type = "p", main = "type = 'p'")
plot(x_line, y_line, type = "b", main = "type = 'b'")
plot(x_line,         type = "h", main = "type = 'h'")
```

## Adding data: ToothGrowth dataset

ToothGrowth: effect of two supplements on tooth growth in 60 guinea pigs.

- `len` — tooth length (numeric)
- `supp` — supplement type: VC or OJ (factor)
- `dose` — dose in milligrams/day (numeric)

```{r}
#| echo: false
#| include: true
library(knitr)
data(ToothGrowth)
kable(head(ToothGrowth))
```

## Adding data: simple plot

```{r}
#| echo: true
#| include: true
data(ToothGrowth)
plot(ToothGrowth$dose, ToothGrowth$len,
     main = "Tooth Growth",
     xlab = "Dose (mg/day)",
     ylab = "Tooth length",
     pch = 19,
     col = "blue")
```

## Adding data: plotting by group

```{r}
#| echo: true
#| include: true
tg_vc <- ToothGrowth[ToothGrowth$supp == "VC", ]
tg_oj <- ToothGrowth[ToothGrowth$supp == "OJ", ]

plot(tg_vc$dose, tg_vc$len,
     main = "Tooth Growth by Supplement Type",
     xlab = "Dose (mg/day)", ylab = "Tooth length",
     pch = 19, col = "blue")
points(tg_oj$dose, tg_oj$len, pch = 19, col = "red")
legend("bottomright", legend = c("VC", "OJ"),
       col = c("blue", "red"), pch = 19)
```

## `plot`, `points`, and `lines`

- `plot()` — creates a new plot
- `points()` — adds points to an existing plot
- `lines()` — adds lines to an existing plot (`points(type = "l")`)

```{r}
#| echo: false
#| include: true
par(mfrow = c(1, 3))
xx <- 1:10
yy <- c(1, 3, 2, 5, 4, 7, 6, 9, 8, 10)
plot(xx, yy, type = "b", main = "single plot")
plot(xx, yy, type = "b", main = "plot + points")
points(xx, yy / 2, type = "p", col = "red")
plot(xx, yy, type = "b", main = "plot + points + lines")
points(xx, yy / 2, type = "l", col = "red")
lines(xx, yy * 2, col = "blue")
```

## Plotting categorical data: `boxplot`

`len ~ supp` is a **formula**: plot `len` on y-axis grouped by `supp`.

```{r}
#| echo: true
#| include: true
boxplot(len ~ supp, data = ToothGrowth,
        main = "Tooth Growth by Supplement Type",
        xlab = "Supplement Type",
        ylab = "Tooth length")
```

## `boxplot` input formats

`boxplot` accepts a formula, a matrix, data frame, or a list.

```{r}
#| echo: false
#| include: true
example_list <- list(
  a = rnorm(100),
  b = rnorm(100, 3, 0.5),
  c = rnorm(100, -2, 5)
)
boxplot(example_list)
```

## Categorical data: `stripchart`

One-dimensional scatter plot — similar use case as `boxplot`.

```{r}
#| echo: true
#| include: true
stripchart(len ~ supp, data = ToothGrowth,
           main = "Tooth Growth by Supplement Type",
           xlab = "Supplement Type",
           ylab = "Tooth length",
           col = "blue",
           vertical = TRUE)
```

## `stripchart` with jitter

When points overlap, use `method = "jitter"` to spread them out.

```{r}
#| echo: true
#| include: true
stripchart(len ~ supp, data = ToothGrowth,
           main = "Tooth Growth by Supplement Type",
           xlab = "Supplement Type",
           ylab = "Tooth length",
           col = "blue",
           method = "jitter",
           vertical = TRUE)
```

## Bar plot vs. Box plot

- **Boxplot**: shows full distribution — median, quartiles, outliers
- **Bar chart**: shows counts or summary values per group

```{r}
#| echo: false
#| include: true
par(mfrow = c(1, 2))
boxplot(mpg ~ cyl, data = mtcars,
        main = "Boxplot of MPG by Cylinders",
        xlab = "Number of Cylinders",
        ylab = "Miles per Gallon",
        col = "lightblue")
barplot(table(mtcars$cyl),
        main = "Bar Chart of Cylinders",
        xlab = "Number of Cylinders",
        ylab = "Frequency",
        col = "lightgreen")
```

## Bar plot example

```{r}
#| echo: true
#| include: true
par(mfrow = c(1, 2))
barplot(table(mtcars$gear),
        main = "Bar Chart of Gears",
        xlab = "Number of Gears", ylab = "Frequency",
        col = "lightgreen")

counts <- table(mtcars$am, mtcars$gear)
barplot(counts,
        main = "Gears by Transmission",
        xlab = "Number of Gears", ylab = "Frequency",
        col = c("lightblue", "lightgreen"))
```

## Histogram

`hist()` takes a numeric vector and computes a histogram.

```{r}
#| echo: true
#| include: true
x2 <- rnbinom(100, mu = 10, size = 10)
hist(x2, main = "Histogram of Negative Binomial Distribution")
```

## Histogram: controlling breaks

`breaks` controls the number of bins.

```{r}
#| echo: true
#| include: true
par(mfrow = c(1, 2))
hist(x2, main = "breaks = 20", breaks = 20)
hist(x2, main = "breaks = 0:40", breaks = 0:40)
```

## Customizing plots: colors, shapes, size

```{r}
#| echo: true
#| include: true
par(mfrow = c(2, 3))
plot(x, y, main = "default plot")
plot(x, y, col = "#FF0000",   main = "col sets point color")
plot(1:20, col = rainbow(20), main = "col as a color vector")
plot(1:20, pch = 1:20,        main = "pch sets point shape")
plot(x, y, cex = 3,           main = "cex sets point size")
plot(x, y, main = "lines and grid")
abline(0, 2)
grid()
```

## Customizing plots: axis limits and log scale

```{r}
#| echo: true
#| include: true
par(mfrow = c(2, 2))
plot(x, y, xlim = c(0, 30), ylim = c(0, 20), type = "b",
     main = "setting axis limits")
plot(x, y, log = "y", type = "b", main = "log scale on y-axis")
plot(x, y, log = "x", type = "b", main = "log scale on x-axis")
plot(x, y, log = "xy", type = "b", main = "log scale on both axes")
```

## Adding legends and annotations

```{r}
#| echo: true
#| include: true
plot(x, y, pch = 19, col = "blue")
points(x + 0.5, y + 0.5, col = "red", pch = 17)
legend("topleft",
       legend = c("Group 1", "Group 2"),
       col = c("blue", "red"),
       pch = c(19, 17))
mtext(expression(alpha + beta == gamma), side = 3, cex = 2)
```

## Combining plots: `par(mfrow)`

Regular grid layout using `par(mfrow = c(rows, cols))`.

```{r}
#| echo: true
#| include: true
par(mfrow = c(2, 3))
plot(x, y, main = "plot 1", col = "1")
plot(x, y, main = "plot 2", col = "2")
plot(x, y, main = "plot 3", col = "3")
plot(x, y, main = "plot 4", col = "4")
plot(x, y, main = "plot 5", col = "5")
plot(x, y, main = "plot 6", col = "6")
```

## Combining plots: `layout`

Custom grid layout — specify cell positions, widths, and heights.

```{r}
#| echo: true
#| include: true
layout(matrix(c(1, 2, 4, 3), nrow = 2),
       width = c(2, 1), height = c(1, 2))
layout.show(4)
```

## Combining plots: `layout` example

```{r}
#| echo: true
#| include: true
layout(matrix(c(1, 2, 4, 3), nrow = 2),
       width = c(2, 1), height = c(1, 2))
hist(x2, main = "1st plot")
plot(x, y, main = "2nd plot")
hist(rnorm(100), main = "3rd plot")
```

## Exporting plots: bitmap formats

`png()` / `jpeg()` / `bmp()` / `tiff()` save a single plot.

```{r}
#| echo: true
#| include: true
dir.create("outputs", showWarnings = FALSE)
png("outputs/plot1.png", width = 800, height = 600)
plot(x, y, main = "Scatter Plot", col = "blue")
dev.off()
```

## Exporting plots: vector formats

`pdf()` and `svg()` produce vector graphics — each `plot()` call adds a new page in a PDF.

```{r}
#| echo: true
#| include: true
pdf("outputs/plot2.pdf", width = 8, height = 6)
plot(x, y, main = "Scatter Plot", col = "blue")
hist(x2, main = "Histogram")
plot(x, y, main = "Line Plot", col = "red")
dev.off()
```

## `par` function

Sets graphical parameters that apply to subsequent plots.

- `par(mfrow = c(2, 2))` — 2×2 grid layout
- `par(mar = c(5, 4, 4, 2) + 0.1)` — plot margins
- `par(cex = 1.5)` — text/symbol magnification
- `par(lty = 2)` — line type (dashed)

See `?par` for the full list of parameters.

## Summary

| Function | Purpose |
|---|---|
| `plot(x, y)` | scatter / line plot |
| `hist(x)` | histogram |
| `boxplot(y ~ group)` | box plot |
| `barplot(counts)` | bar chart |
| `stripchart(y ~ group)` | strip chart |
| `points()` / `lines()` | add to existing plot |
| `legend()` | add legend |
| `par()` | graphical parameters |
| `png()` / `pdf()` | export to file |