---
title: "Analysis of Differential Gene Expression with DESeq2"
format:
  revealjs:
    self-contained: true
    slide-number: true
editor: visual
---
<style>
.reveal {
  font-size: 160%;
}
</style>

```{r}
#| echo: false
#| include: false
library(DESeq2)
library(ggplot2)
library(plotly)
library(matrixStats)
library(pheatmap)
library(heatmaply)
library(htmlwidgets)
library(org.Dm.eg.db)
library(AnnotationDbi)
library(knitr)
library(kableExtra)
```

## Introduction
This presentation introduces a complete RNA-Seq analysis workflow using the **DESeq2**  package. You will learn how to:

- Load and quality-filter RNA-Seq count data.
- Perform differential expression analysis.
- Explore the data with PCA, MDS, and pairs plots.
- Visualize results interactively using **plotly**  and **heatmaply** .
- Annotate genes with FlyBase information.

The pasilla dataset comes from an experiment investigating the effect of siRNA knockdown of the *pasilla* gene in Drosophila.


## Setup and Load Libraries
```{r}
#| echo: true
#| include: true
#| eval: false
library(DESeq2)       # Differential expression analysis
library(ggplot2)      # Static plotting
library(plotly)       # Interactive plotting
library(matrixStats)  # For computing row variances
library(pheatmap)     # Static heatmaps
library(heatmaply)    # Interactive heatmaps
library(htmlwidgets)  # Saving interactive widgets
library(org.Dm.eg.db) # Drosophila gene annotations - Must be installed with constraint `BiocManager::install("org.Dm.eg.db")`
library(AnnotationDbi)# Annotation database interface



```

## Data Loading — Count Matrix
```{r}
#| echo: true
#| include: true
# Load the count matrix and sample annotation.
# Adjust file paths as needed.
countData <- read.delim("../data/pasilla_gene_counts.tsv", row.names = 1)
head(countData)
```

## Data Loading — Sample Annotation
```{r}
#| echo: true
#| include: true
colData <- read.csv("../data/pasilla_sample_annotation.csv", row.names = 1)
# Ensure the 'condition' variable is a factor.
colData$condition <- as.factor(colData$condition)
head(colData)
```

## Align Sample Order
```{r}
#| echo: true
#| include: true
# It is critical that the order of columns in countData matches the rows in colData.
all(rownames(colData) == colnames(countData))  # Should be TRUE
colData <- colData[colnames(countData), ]
all(rownames(colData) == colnames(countData))
print(colData)
```

## Create DESeqDataSet
```{r}
#| echo: true
#| include: true
# Create a DESeqDataSet object using the count data and sample annotation.
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ condition)

# Add gene names as metadata to the DESeqDataSet.
metadata <- data.frame(gene = rownames(countData))
mcols(dds) <- DataFrame(mcols(dds), metadata)
print(dds)
```

## Filter Genes with Low Counts
```{r}
#| echo: true
#| include: true
# Filter out genes with low counts (total count < 3) to reduce noise.
keep <- rowSums(counts(dds)) >= 3
dds <- dds[keep, ]
cat("Number of genes after filtering:", nrow(dds), "\n")
# Set 'untreated' as the reference level for condition.
print(levels(dds$condition))
dds$condition <- relevel(dds$condition, ref = "untreated")
print(levels(dds$condition))
```

> 📝 **Task 1** — Plot a histogram of the row sums to examine the count distribution before filtering.

> 📝 **Task 2** — Use `table()` to determine how many genes have total counts less than 3.

## Differential Expression Analysis
```{r}
#| echo: true
#| include: true
# Run the full DESeq2 pipeline:
# 1. Normalization, 2. Dispersion estimation, 3. Model fitting, and 4. Hypothesis testing.
dds <- DESeq(dds)

# Extract results with the default FDR threshold (0.1).
res <- results(dds)
summary(res)
```

## DESeq2 Results
```{r}
#| echo: true
#| include: true
# Extract results using a stricter FDR threshold (alpha = 0.05).
res05 <- results(dds, alpha = 0.05)
summary(res05)
```

## Variance Stabilizing Transformation (VST)
```{r}
#| echo: true
#| include: true
# RNA-Seq count data have wide ranges and variance increases with the mean.
# VST transforms the data so that variance is approximately constant, making it
# better suited for visualizations like PCA and heatmaps.
vsd <- vst(dds, blind = FALSE)
pcaData <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
pcaData$type <- colData$type  # Add sequencing type for extra context.
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

## PCA: Principal Component Analysis
```{r}
#| echo: true
#| include: true
p_pca <- ggplot(pcaData, aes(PC1, PC2, color = condition, shape = type)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of pasilla Samples") +
  theme_minimal()
p_pca
```

## Interactive PCA Plot
```{r}
#| echo: true
#| include: true
# Convert the PCA plot into an interactive plot using plotly.
ggplotly(p_pca)
```

## MDS: Multidimensional Scaling
```{r}
#| echo: true
#| include: true
# Compute a distance matrix from the VST data and perform MDS.
dists <- dist(t(assay(vsd)))
mds <- cmdscale(dists)
mdsData <- data.frame(MDS1 = mds[, 1],
                      MDS2 = mds[, 2],
                      condition = colData$condition,
                      type = colData$type)
# Static MDS plot with ggplot2.
p_mds <- ggplot(mdsData, aes(MDS1, MDS2, color = condition, shape = type)) +
  geom_point(size = 3) +
  ggtitle("MDS Plot of pasilla Samples") +
  theme_minimal()
p_mds
```

## Pairs Plot
```{r}
#| echo: true
#| include: true
# Create a pairs plot to explore relationships between samples.
# The transparency (alpha) is set to reduce overplotting.
vsd_df <- as.data.frame(assay(vsd))
pairs(vsd_df, col="#00000010")
```

## MA Plot
```{r}
#| echo: true
#| include: true
# The MA plot shows the log2 fold changes versus the mean expression.
# Significant genes (FDR < 0.05) are highlighted.
plotMA(res05, main = "DESeq2 MA-Plot (FDR < 0.05)", ylim = c(-5, 5))
```

## Volcano Plot
```{r}
#| echo: true
#| include: true
# Create a volcano plot to display the relationship between fold change and significance.
volcanoData <- as.data.frame(res)
volcanoData$gene <- rownames(volcanoData)
p_volcano <- ggplot(volcanoData, aes(x = log2FoldChange, y = -log10(padj), text = gene)) +
  geom_point(alpha = 0.4) +
  ggtitle("Volcano Plot") +
  xlab("Log2 Fold Change") +
  ylab("-Log10 Adjusted P-Value") +
  theme_minimal()
p_volcano
```

## Interactive Volcano Plot
```{r}
#| echo: true
#| include: true
# Make the volcano plot interactive.
ggplotly(p_volcano, tooltip = c("text", "x", "y"))
```

## Accounting for Technical Variation
Sometimes technical variables—like sequencing type (single-read vs. paired-end)—can affect your results. To account for this, include the technical variable as a covariate in your design:

```{r}
#| echo: true
#| include: true
# Adjust the 'type' variable to remove extra characters and ensure it's a factor.
colData$type <- as.factor(gsub("-.+","", colData$type))
dds2 <- DESeqDataSetFromMatrix(countData = countData,
                               colData = colData,
                               design = ~ type + condition)
dds2$condition <- relevel(dds2$condition, ref = "untreated")
dds2 <- DESeq(dds2)

# Extract results accounting for technical variation.
res2 <- results(dds2, contrast = c("condition", "untreated", "treated"))
summary(res2)
```

> 📝 **Task 3** — Compare the number of DEGs (FDR < 0.05) from this model with the previous one.

## Heatmap: Top Expressed Genes
```{r}
#| echo: true
#| include: true
# Normalize data for heatmap visualization.
ntd2 <- normTransform(dds2)
df <- as.data.frame(colData(dds2)[, c("condition", "type")])
# Heatmap of the top 20 most expressed genes.
select <- order(rowMeans(counts(dds2, normalized = TRUE)), decreasing = TRUE)[1:20]
pheatmap(assay(ntd2)[select, ], cluster_rows = FALSE, show_rownames = FALSE,
         cluster_cols = FALSE, annotation_col = df)
```

## Heatmap: Differentially Expressed Genes
```{r}
#| echo: true
#| include: true
# Heatmap with hierarchical clustering for differentially expressed genes (padj < 0.01).
res2_05 <- results(dds2, alpha = 0.05, contrast = c("condition", "untreated", "treated"))
select <- res2_05$padj < 0.01
select[is.na(select)] <- FALSE
pheatmap(assay(ntd2)[select, ], cluster_rows = TRUE, show_rownames = FALSE,
         cluster_cols = TRUE, annotation_col = df, scale = "row",
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

> 📝 **Task 4** — Export this heatmap as a PNG file using `png()`, `pheatmap()`, and `dev.off()`.

## Interactive Heatmap with heatmaply
```{r}
#| echo: true
#| include: true
# Create an interactive heatmap and export it as an HTML file.
p <- heatmaply(assay(ntd2)[select, ], scale = "row", k_row = 3, k_col = 3)
saveWidget(p, "../work_dir/heatmaply.html")
p  # Display interactive heatmap
```

## Gene Annotation
```{r}
#| echo: true
#| include: true
# Retrieve gene annotations using the org.Dm.eg.db package.
# This provides gene symbols and descriptions based on FlyBase IDs.
fly_ids <- rownames(countData)
fly_genes <- select(org.Dm.eg.db, keys = fly_ids, columns = c("SYMBOL", "GENENAME"),
                    keytype = "FLYBASE")

# Merge annotation data with differential expression results.
res2_05_reordered <- res2_05[order(res2_05$padj), ]
res2_05_annotated <- merge(as.data.frame(res2_05_reordered),
                           fly_genes, by.x = "row.names",
                           by.y = "FLYBASE", all.x = TRUE)
res2_05_annotated <- res2_05_annotated[order(res2_05_annotated$padj), ]
write.csv(res2_05_annotated, "../work_dir/deseq2_results_annotated.csv", row.names = FALSE)
```

> 📝 **Task 5** — Create a data frame combining `res2_05_annotated` with per-sample expression values from `assay(vsd2)`.

## Annotated Results Table
```{r}
#| echo: false
#| include: true
kable(head(res2_05_annotated, 20), format.args = list(scientific = TRUE, digits = 2)) %>%
  kable_styling(font_size = 14, full_width = FALSE)
```
