---
title: "GenomicRanges: Bioconductor Package for Genomic Data"
format:
  revealjs:
    self-contained: true
    slide-number: true
editor: visual
---

<style>
.reveal {
  font-size: 160%;
}
</style>

```{r}
#| echo: false
#| include: false
library(GenomicRanges)
library(Biostrings)
library(Gviz)

# Helper used throughout the lesson — explained on its own slide below.
plotGenomicTracks <- function(tracks_list,
                              colors     = NULL,
                              chromosome = NULL,
                              from       = NULL,
                              to         = NULL) {
  n <- length(tracks_list)
  if (is.null(names(tracks_list)) || any(names(tracks_list) == "")) {
    stop("tracks_list must be a named list (names used as track titles).")
  }
  if (!is.null(colors) && length(colors) != n) {
    stop("If colors is provided, its length must equal number of tracks.")
  }
  if (is.null(colors)) colors <- rep("gray", n)

  annotation_tracks <- mapply(function(gr, col, nm) {
    args <- list(range = gr, name = nm, fill = col,
                 col = "black", stacking = "dense")
    if ("id" %in% names(mcols(gr))) {
      ids <- mcols(gr)$id
      if (!all(is.na(ids)) && length(ids) == length(gr)) {
        args$showFeatureId     <- TRUE
        args$featureAnnotation <- "id"
        args$fontcolor.feature <- "black"
      }
    }
    do.call(AnnotationTrack, args)
  },
  gr = tracks_list, col = colors, nm = names(tracks_list),
  SIMPLIFY = FALSE)

  axis_track <- GenomeAxisTrack()
  all_tracks <- c(list(axis_track), annotation_tracks)
  plotTracks(all_tracks,
             chromosome       = chromosome,
             from             = from,
             to               = to,
             background.panel = "#FFFEDB",
             col.axis         = "black",
             cex.title        = 0.8,
             extend.left      = 5,
             extend.right     = 5)
  invisible(annotation_tracks)
}
```

# Part 1 — What are genomic ranges?

## What are genomic ranges?

A genomic range = an interval on a genome.

It is defined by:

-   **seqname** — chromosome or contig (e.g. `"chr1"`)
-   **start / end** — 1-based, end-inclusive
-   **strand** — `+`, `-`, or `*`
-   optional **metadata** — gene name, score, GC, ...

Used for: genes, exons, peaks, variants, primer hits.

## Core classes in GenomicRanges

-   **`GRanges`** — a vector of intervals (the workhorse).
-   **`GRangesList`** — a list of GRanges (e.g., exons grouped by transcript).
-   **`GPos`** — single-position version (rarely used here).

## Relationship with Biostrings

-   `Biostrings` handles **sequence content** (DNAStringSet, ...).
-   `GenomicRanges` handles **coordinates**.
-   Together: extract the DNA at an interval, count GC, scan for motifs.

# Part 2 — Plotting ranges

## Why plot ranges?

Range operations are easier to **see** than to read off a printed table.

Throughout this lesson we use a helper, `plotGenomicTracks()`, that wraps
**Gviz** — Bioconductor's track-plotting package. We will build on Gviz
in later lessons, so it is worth meeting it here.

## The `plotGenomicTracks()` helper

```{r}
#| echo: true
#| include: true
#| eval: false
plotGenomicTracks <- function(tracks_list, colors = NULL,
                              chromosome = NULL,
                              from = NULL, to = NULL) {
  n <- length(tracks_list)
  if (is.null(colors)) colors <- rep("gray", n)
  annotation_tracks <- mapply(function(gr, col, nm) {
    args <- list(range = gr, name = nm, fill = col,
                 col = "black", stacking = "dense")
    if ("id" %in% names(mcols(gr))) {
      args$showFeatureId     <- TRUE
      args$featureAnnotation <- "id"
    }
    do.call(AnnotationTrack, args)
  }, gr = tracks_list, col = colors, nm = names(tracks_list),
     SIMPLIFY = FALSE)
  plotTracks(c(list(GenomeAxisTrack()), annotation_tracks),
             chromosome = chromosome, from = from, to = to,
             background.panel = "#FFFEDB", cex.title = 0.8)
}
```

Each named GRanges becomes one Gviz `AnnotationTrack`.

## How to call `plotGenomicTracks()`

-   **Input** — a *named* list of `GRanges` objects.
    -   Names → track titles (left side of plot).
-   **Output** — a Gviz multi-track figure with a genome axis on top.
-   Add an `id` metadata column to label individual intervals.

```{r}
#| echo: true
#| include: true
#| eval: false
plotGenomicTracks(list(gr1 = gr1, gr2 = gr2))
```

## Trying `plotGenomicTracks()`

```{r}
#| echo: true
#| include: true
#| fig-height: 2.5
demo <- GRanges("chr1",
                IRanges(start = c(5, 30, 60), end = c(20, 45, 80)),
                strand = c("+", "-", "*"),
                id = c("a", "b", "c"))
plotGenomicTracks(list(demo = demo))
```

# Part 3 — Creating and inspecting GRanges

## Creating a GRanges object

```{r}
#| echo: true
#| include: true
gr <- GRanges(
  seqnames = c("chr1", "chr2", "chr2", "chr1", "chr3"),
  ranges   = IRanges(start = c(10, 15, 20, 25, 30),
                     end   = c(20, 25, 30, 35, 40)),
  strand   = c("+", "-", "+", "*", "-"),
  score    = c(5, 10, 15, 20, 25),
  GC       = c(0.45, 0.55, 0.60, 0.70, 0.80)
)
gr
```

## Visualizing our GRanges

Gviz draws one chromosome at a time — show only the chr1 ranges.

```{r}
#| echo: true
#| include: true
#| fig-height: 2.5
plotGenomicTracks(list(gr_chr1 = gr[seqnames(gr) == "chr1"]))
```

## Accessing GRanges components

```{r}
#| echo: true
#| include: true
seqnames(gr)
strand(gr)
gr$score
width(gr)
length(gr)
```

> 📝 **Task 1** — create a GRanges of three intervals on `"chr1"` with strands
> of your choice and a metadata column `name`. Print and plot it.

# Part 4 — Subsetting GRanges

## Subsetting GRanges

```{r}
#| echo: true
#| include: true
gr[1:3]                                          # by index
gr[width(gr) > 10]                               # by width
gr[seqnames(gr) == "chr1"]                       # by chromosome
```
## Subsetting GRanges
```{r}
#| echo: true
#| include: true
gr[strand(gr) == "+"]                            # by strand
gr[gr$score > 10]                                # by metadata
gr[seqnames(gr) == "chr2" & strand(gr) == "+"]   # combined
```

> 📝 **Task 2** — from `gr`, keep only ranges on chr1 OR chr2 with
> `score >= 10`. How many ranges remain?

# Part 5 — Basic range operations

## A toy GRanges to play with

```{r}
#| echo: true
#| include: true
g <- GRanges("chr1",
             IRanges(start = c(10, 30, 60), end = c(20, 45, 75)),
             strand = c("+", "-", "+"))
```

We will visualize each range operation against this same `g`.

## `shift()` — move every range

```{r}
#| echo: true
#| include: true
#| fig-height: 2.5
plotGenomicTracks(list(original = g, shifted = shift(g, 5)))
```

## `resize()` — fixed width, choose anchor

```{r}
#| echo: true
#| include: true
#| fig-height: 3.5
plotGenomicTracks(list(
  original     = g,
  resize_start = resize(g, width = 10, fix = "start"),
  resize_end   = resize(g, width = 10, fix = "end")
))
```

`fix` controls what stays put: `"start"`, `"end"`, or `"center"`.

## `narrow()` — trim from either end

```{r}
#| echo: true
#| include: true
#| fig-height: 2.5
plotGenomicTracks(list(original = g,
                 narrow   = narrow(g, start = 2, end = -2)))
```

`start = 2` drops 1 bp from the start; `end = -2` drops 1 bp from the end.

## `flank()` — adjacent windows

Useful for promoters (5′ flank) and 3′ UTRs.

```{r}
#| echo: true
#| include: true
#| fig-height: 3.5
plotGenomicTracks(list(
  original     = g,
  flank_5prime = flank(g, width = 5, start = TRUE),
  flank_3prime = flank(g, width = 5, start = FALSE)
))
```

## `restrict()` — clip to a window

```{r}
#| echo: true
#| include: true
#| fig-height: 2.5
plotGenomicTracks(list(original   = g,
                 restricted = restrict(g, start = 15, end = 50)))
```

> 📝 **Task 3** — build a GRanges of 3 "genes" on chr1, then extract a
> **200 bp promoter** for each (use `flank()` with `start = TRUE`). Plot
> both tracks together.

# Part 6 — Finding overlaps

## Why overlap operations?

The single most common operation in genomics:

-   *"Which of my peaks fall inside a gene?"*
-   *"Which variants hit an exon?"*
-   *"How many reads cover this region?"*

All use the same overlap machinery.

## `findOverlaps()`

```{r}
#| echo: true
#| include: true
#| fig-height: 2
gr1 <- GRanges("chr1", IRanges(start = c(5, 80, 130), width = 20))
gr2 <- GRanges("chr1", IRanges(start = c(10, 100, 125), width = 15))
plotGenomicTracks(list(gr1 = gr1, gr2 = gr2))
```

```{r}
#| echo: true
#| include: true
hits <- findOverlaps(gr1, gr2)
hits
```

## Reading a `Hits` object

```{r}
#| echo: true
#| include: true
queryHits(hits)     # indices into gr1
subjectHits(hits)   # indices into gr2
```

Each pair `(queryHits[i], subjectHits[i])` is one overlapping pair.

## `countOverlaps()`

```{r}
#| echo: true
#| include: true
countOverlaps(gr1, gr2)
```

For each range in `gr1`, how many ranges in `gr2` it overlaps.

## `subsetByOverlaps()`

```{r}
#| echo: true
#| include: true
#| fig-height: 3
gr1_overlapping <- subsetByOverlaps(gr1, gr2)
plotGenomicTracks(list(gr1 = gr1, gr2 = gr2,
                 gr1_overlapping = gr1_overlapping))
```

## `overlapsAny()`

```{r}
#| echo: true
#| include: true
overlapsAny(gr1, gr2)
```

A logical vector — `TRUE` where the range has at least one overlap.

> 📝 **Task 4** — build two GRanges on chr1 of your choice and use
> `subsetByOverlaps()` to keep only the ranges of the first that overlap any
> range of the second. Plot input, reference, and result.

# Part 7 — Nearest-neighbor operations

## `nearest()` and `distance()`

```{r}
#| echo: true
#| include: true
#| fig-height: 2
query   <- GRanges("chr1", IRanges(c(10, 25, 55), width = 5))
subject <- GRanges("chr1", IRanges(c(1, 18, 40, 70), width = 5))
plotGenomicTracks(list(query = query, subject = subject))
```

```{r}
#| echo: true
#| include: true
nearest(query, subject)
distance(query, subject[1:3])
```

## `precede()` and `follow()` — what they answer

Unlike `nearest()`, which can return a subject that *overlaps* the query,
`precede` and `follow` always look **outside** the query and pick a
**direction**.

-   `precede(x, s)` — for each `x`, the index of the **first range in `s`
    that comes *after* `x` along the 5′ → 3′ direction**.
    *(In the printed gene-track sense: the next downstream feature.)*
-   `follow(x, s)` — for each `x`, the index of the **last range in `s` that
    came *before* `x` along the 5′ → 3′ direction**.
    *(The previous upstream feature.)*

If no such range exists, the result is `NA`.

## Why "precede" feels backwards at first

Read the names from the **subject's** point of view:

-   *"Which range in `s` does `x` **precede**?"* → the one **after** `x`.
-   *"Which range in `s` does `x` **follow**?"*  → the one **before** `x`.

So `precede(x, s)` returns the **downstream** neighbor of `x`, and
`follow(x, s)` returns the **upstream** neighbor of `x`. Many people
remember it the other way around — keep this slide handy.

## A worked example

```{r}
#| echo: true
#| include: true
#| fig-height: 2.2
q <- GRanges("chr1", IRanges(c(25), width = 5), strand = "+", id = "q")
s <- GRanges("chr1", IRanges(c(5, 15, 50, 70), width = 5),
             strand = "+", id = paste0("s", 1:4))
plotGenomicTracks(list(query = q, subject = s))
```

```{r}
#| echo: true
#| include: true
precede(q, s)   # which subject comes AFTER q?  → s3 (index 3)
follow(q, s)    # which subject came BEFORE q?  → s2 (index 2)
```

So `q` *precedes* `s[3]` and *follows* `s[2]`.

## Distance to the precede/follow neighbor

A common pattern: find the upstream/downstream neighbor and how far away it
is.

```{r}
#| echo: true
#| include: true
i_down <- precede(q, s)
i_up   <- follow(q, s)

distance(q, s[i_down])   # bp to next downstream feature
distance(q, s[i_up])     # bp to previous upstream feature
```

This is exactly how you'd annotate "distance to next gene" for ChIP-seq
peaks or variants.

## Why the strand matters

The 5′ → 3′ direction depends on **strand**, but plot positions go
left-to-right. So on a `-`-strand gene, the *upstream* neighbor is on the
**right** of the plot, and `follow()` returns it.

```{r}
#| echo: true
#| include: true
q_minus <- GRanges("chr1", IRanges(25, width = 5), strand = "-", id = "q-")
s_plus  <- GRanges("chr1", IRanges(c(5, 50), width = 5),
                   strand = "+", id = c("s1", "s2"))
precede(q_minus, s_plus)   # NA: opposite strands
```

Mixed strands often return `NA` — see the table on the next slide.

## Strand effects on precede / follow {.smaller}

The strand of `x` and `subject` determines whether a comparison is even
defined. Set `ignore.strand = TRUE` to disable this check.

|  x  | subject | orientation                                          |
|:---:|:-------:|:-----------------------------------------------------|
|  +  |    +    | --->                                                 |
|  +  |    -    | NA                                                   |
|  +  |    \*   | --->                                                 |
|  -  |    -    | <---                                                 |
|  -  |    \*   | <---                                                 |
|  \* |    +    | --->                                                 |
|  \* |    \*   | --->  (only case where `*` arbitrarily means `+`)    |

The single exception: when **both** `x` and `subject` are `*`, both are
treated as `+`.

# Part 8 — Set operations on ranges

## Setup

```{r}
#| echo: true
#| include: true
gr_a <- GRanges("chr1", IRanges(start = c(10, 20, 30), end = c(15, 25, 35)))
gr_b <- GRanges("chr1", IRanges(start = c(12, 22, 40), end = c(17, 27, 45)))
```

## `union()`

```{r}
#| echo: true
#| include: true
#| fig-height: 3
plotGenomicTracks(list(gr_a = gr_a, gr_b = gr_b,
                 union = union(gr_a, gr_b)))
```

## `intersect()`

```{r}
#| echo: true
#| include: true
#| fig-height: 3
plotGenomicTracks(list(gr_a = gr_a, gr_b = gr_b,
                 intersect = intersect(gr_a, gr_b)))
```

## `setdiff()`

```{r}
#| echo: true
#| include: true
#| fig-height: 3
plotGenomicTracks(list(gr_a = gr_a, gr_b = gr_b,
                 setdiff = setdiff(gr_a, gr_b)))
```

> 📝 **Task 5** — using `gr_a` and `gr_b` above, compute the **symmetric
> difference** (parts in either but not in both).
> Hint: `setdiff(union(...), intersect(...))`.

# Part 9 — GRangesList

## Why GRangesList?

Use a `GRangesList` when intervals come in groups:

-   exons of a transcript,
-   peaks per sample,
-   binding sites per TF.

It is a list whose elements are all `GRanges`.

## Creating a GRangesList

```{r}
#| echo: true
#| include: true
grl <- GRangesList(
  sample1 = GRanges(seqnames = c("chr1", "chr2"),
                    ranges   = IRanges(start = c(10, 20), end = c(20, 30))),
  sample2 = GRanges(seqnames = c("chr1", "chr3"),
                    ranges   = IRanges(start = c(15, 40), end = c(25, 50)))
)
grl
```

## Working with a GRangesList

```{r}
#| echo: true
#| include: true
grl[[1]]              # first element
grl$sample2           # by name
elementNROWS(grl)     # length of each element
unlist(grl)           # flatten
```

`endoapply(grl, shift, 5)` applies `shift()` to each element and returns a
GRangesList of the same shape.

# Part 10 — Integration with Biostrings

## Sequences from coordinates

`getSeq()` extracts DNA from a `BSgenome` at the given GRanges coordinates.

```{r}
#| echo: true
#| include: true
#| eval: false
library(BSgenome.Hsapiens.UCSC.hg38)

genes <- GRanges(
  seqnames = c("chr1", "chr2"),
  ranges   = IRanges(start = c(10000, 20000), end = c(10500, 20500)),
  strand   = c("+", "-"),
  gene_id  = c("gene1", "gene2")
)

gene_seqs <- getSeq(BSgenome.Hsapiens.UCSC.hg38, genes)
names(gene_seqs) <- genes$gene_id
gene_seqs
letterFrequency(gene_seqs, "GC", as.prob = TRUE)
```

The BSgenome package is large (~1 GB) — skip if not installed.
