---
title: "Hands-On Data Cleaning & Manipulation with Tidyverse"
subtitle: "Golub et al. Leukemia Gene Expression Data"
format:
  revealjs:
    self-contained: true
    slide-number: true
editor: visual
---
<style>
.reveal {
  font-size: 160%;
}
</style>

```{r}
#| echo: false
#| include: false
knitr::opts_knit$set(root.dir = "../../")
library(tidyverse)
library(pheatmap)
library(knitr)
```

## Overview

This walkthrough applies tidyverse tools to a real microarray dataset.

You will learn to:

- Load and inspect gene expression data with `readr`
- Clean columns with `select()` and `filter()`
- Add new variables with `mutate()`
- Summarize groups with `group_by()` / `summarize()`
- Reshape data with `pivot_longer()` / `pivot_wider()`
- Merge data frames with `left_join()`
- Visualize results with `ggplot2` and `pheatmap`

## The Dataset

**Golub et al. (1999)** — gene expression profiling of two types of acute leukemia:

- **ALL** — Acute Lymphoblastic Leukemia
- **AML** — Acute Myeloid Leukemia

Data files in `data/aml/`:

| File | Description |
|---|---|
| `data_set_ALL_AML_train.csv` | Expression matrix + detection calls |
| `annotation.csv` | Patient → cancer type mapping |

## 1. Load the Data

```{r}
#| echo: true
#| include: true
library(tidyverse)
train_file <- "data/aml/data_set_ALL_AML_train.csv"
expr <- read_csv(train_file)
```

## 1. Inspect the Data

```{r}
#| echo: true
#| include: true
print(dim(expr))
print(colnames(expr)[1:10])
```

## 1. Inspect the Data (cont'd)

```{r}
#| echo: true
#| include: true
print(head(expr[, 1:5]))
```

Each row is a gene. Numeric columns are expression values; `call*` columns are detection flags (`"P"`, `"M"`, `"A"`).

## 2. Remove "call" Columns

```{r}
#| echo: true
#| include: true
expr_clean <- expr %>% select(-contains("call"))
print(colnames(expr_clean)[1:10])
```

## 2. Verify Dimensions

```{r}
#| echo: true
#| include: true
print(dim(expr_clean))
```

`-contains("call")` dropped every column whose name includes `"call"`.

## 3. Filter Kinase Genes

```{r}
#| echo: true
#| include: true
kinase_genes <- expr_clean %>%
  filter(str_detect(`Gene Description`, "kinase"))
print(paste("Number of kinase-related genes:", nrow(kinase_genes)))
print(head(kinase_genes$`Gene Description`, 10))
```

## 3. Filter Ribosomal Genes

```{r}
#| echo: true
#| include: true
ribosomal_genes <- expr_clean %>%
  filter(str_detect(`Gene Description`, "ribosomal"))
print(paste("Number of ribosomal-related genes:", nrow(ribosomal_genes)))
```

## 4. Flag Control Probes

Control probes have accession numbers starting with `"AFFX"`. The `^` anchors to the start of the string.

```{r}
#| echo: true
#| include: true
expr_clean <- expr_clean %>%
  mutate(IsControl = str_detect(`Gene Accession Number`, "^AFFX"))
print(head(expr_clean %>%
  select(`Gene Description`, `Gene Accession Number`, IsControl)))
```

## 4. Count Control vs. Biological Probes

```{r}
#| echo: true
#| include: true
print(table(expr_clean$IsControl))
```

## 5. Compute Average Expression per Gene

`rowMeans()` calculates the mean across all sample columns for each row.

```{r}
#| echo: true
#| include: true
expr_clean <- expr_clean %>%
  mutate(AvgExpr = rowMeans(
    select(., -c(`Gene Description`, `Gene Accession Number`, IsControl)),
    na.rm = TRUE
  ))
print(head(expr_clean %>% select(`Gene Description`, AvgExpr)))
```

## 5. Summarize by Control Status

```{r}
#| echo: true
#| include: true
control_summary <- expr_clean %>%
  group_by(IsControl) %>%
  summarize(mean_expr = mean(AvgExpr), sd_expr = sd(AvgExpr))
print(control_summary)
```

Control probes show systematically different expression levels from biological genes.

## 6. Reshape: Wide → Long Format

Long format (one row per gene–sample) is required by `ggplot2`.

```{r}
#| echo: true
#| include: true
expr_long <- expr_clean %>%
  pivot_longer(
    cols = -c(`Gene Description`, `Gene Accession Number`, IsControl, AvgExpr),
    names_to = "Sample",
    values_to = "Expression"
  )
print(head(expr_long))
```

## 6. Long Format Dimensions

```{r}
#| echo: true
#| include: true
print(dim(expr_long))  # rows = genes × samples
```

## 7. Reshape Call Columns

```{r}
#| echo: true
#| include: true
calls_long <- expr %>%
  select(`Gene Accession Number`, contains("call")) %>%
  pivot_longer(
    cols = -`Gene Accession Number`,
    names_to = "SampleCall",
    values_to = "Call"
  ) %>%
  mutate(Sample = str_remove(SampleCall, "^call[.]+")) %>%
  select(-SampleCall)
print(head(calls_long))
```

## 7. Renumber Samples and Join

```{r}
#| echo: true
#| include: true
f <- function(x) { (x - 2) / 2 }
calls_long$Sample <- calls_long$Sample %>%
  as.numeric() %>% f() %>% as.character()

expr_long <- expr_long %>%
  left_join(calls_long, by = c("Gene Accession Number", "Sample"))
print(table(expr_long$Call, useNA = "ifany"))
```

## 8. Boxplot by Call Category

```{r}
#| echo: true
#| include: true
#| fig-height: 4.5
p1 <- ggplot(expr_long, aes(x = Call, y = Expression)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(trans = "log10") +
  labs(title = "Expression by Call Category",
       x = "Call (Detection Flag)", y = "Expression (log10)") +
  theme_minimal()
print(p1)
```

## 8. Top 10 Kinase Genes — Bar Plot

```{r}
#| echo: true
#| include: true
#| fig-height: 5
#| fig-width: 11
top10_kinases <- expr_clean %>%
  filter(str_detect(`Gene Description`, "kinase")) %>%
  arrange(desc(AvgExpr)) %>%
  slice_head(n = 10) %>%
  mutate(label = str_trunc(`Gene Description`, 50))
p2 <- ggplot(top10_kinases,
             aes(x = reorder(label, AvgExpr), y = AvgExpr)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Kinase Genes", x = NULL, y = "Average Expression") +
  theme_minimal(base_size = 14)
print(p2)
```

## 9. Load Annotation

```{r}
#| echo: true
#| include: true
annotation <- read_csv("data/aml/annotation.csv")
annotation <- annotation %>%
  rename(Sample = patient) %>%
  mutate(Sample = as.character(Sample))
print(head(annotation))
```

## 9. Merge Annotation with Expression Data

```{r}
#| echo: true
#| include: true
expr_long <- expr_long %>% left_join(annotation, by = "Sample")
print(head(expr_long %>% select(`Gene Description`, Sample, Expression, cancer)))
```

## 9. Faceted Boxplot — Top 5 Kinase Genes

```{r}
#| echo: true
#| include: true
#| fig-height: 4.5
top5_kinases <- expr_clean %>%
  filter(str_detect(`Gene Description`, "kinase")) %>%
  arrange(desc(AvgExpr)) %>%
  slice_head(n = 5) %>%
  pull(`Gene Description`)
kinase_subset <- expr_long %>%
  filter(`Gene Description` %in% top5_kinases)
p4 <- ggplot(kinase_subset, aes(x = cancer, y = Expression, fill = cancer)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  facet_wrap(~ `Gene Description`) +
  labs(title = "Top 5 Kinase Genes by Cancer Type",
       x = "Cancer Type", y = "Expression") +
  theme_minimal() + theme(legend.position = "none")
print(p4)
```

## 10. Compute Variability Metrics

```{r}
#| echo: true
#| include: true
expr_clean <- expr_clean %>%
  mutate(
    Variance = apply(
      select(., -c(`Gene Description`, `Gene Accession Number`, IsControl, AvgExpr)),
      1, var),
    SD = apply(
      select(., -c(`Gene Description`, `Gene Accession Number`, IsControl, AvgExpr)),
      1, sd),
    CoeffVar = SD / AvgExpr
  )
print(head(expr_clean %>% select(`Gene Description`, AvgExpr, SD, CoeffVar)))
```

**CoeffVar** = SD / mean — normalizes variability by expression level.

## 10. Heatmap — Top 100 Variable Genes {.smaller}

```{r}
#| echo: true
#| include: true
#| fig-height: 4.5
library(pheatmap)
top100_var_expr <- expr_clean %>%
  arrange(desc(CoeffVar)) %>%
  slice_head(n = 100) %>%
  select(-c(`Gene Description`, `Gene Accession Number`,
            IsControl, AvgExpr, Variance, SD, CoeffVar))
pheatmap(top100_var_expr, scale = "row", show_rownames = FALSE,
         cluster_rows = TRUE, cluster_cols = TRUE,
         main = "Top 100 Genes — Highest Variance")
```

## 10. Heatmap with Cancer Type Annotation {.smaller}

```{r}
#| echo: true
#| include: true
#| fig-height: 4.5
sample_cols <- colnames(top100_var_expr)
ann <- data.frame(
  Cancer = annotation$cancer[match(sample_cols, annotation$Sample)],
  row.names = sample_cols
)
pheatmap(top100_var_expr, scale = "row", show_rownames = FALSE,
         cluster_rows = TRUE, cluster_cols = TRUE,
         main = "Top 100 Variable Genes by Cancer Type",
         annotation_col = ann)
```

## 10. Heatmap — Top 100 by Average Expression {.smaller}

```{r}
#| echo: true
#| include: true
#| fig-height: 4
top100_avg <- expr_clean %>%
  arrange(desc(AvgExpr)) %>%
  slice_head(n = 100)
top100_avg_expr <- top100_avg %>%
  select(-c(`Gene Description`, `Gene Accession Number`,
            IsControl, AvgExpr, Variance, SD, CoeffVar))
gene_desc <- top100_avg$`Gene Description`
pheatmap(top100_avg_expr, scale = "row",
         labels_row = gene_desc, show_rownames = TRUE,
         cluster_rows = TRUE, cluster_cols = TRUE,
         clustering_method = "ward.D",
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         main = "Top 100 Genes — Highest Average Expression",
         annotation_col = ann,
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

## Summary

| Step | Function(s) |
|---|---|
| Load data | `read_csv()` |
| Remove columns | `select(-contains(...))` |
| Filter rows | `filter()`, `str_detect()` |
| Add columns | `mutate()`, `rowMeans()` |
| Group summaries | `group_by()`, `summarize()` |
| Reshape | `pivot_longer()`, `pivot_wider()` |
| Merge | `left_join()` |
| Visualize | `ggplot2`, `pheatmap` |
