---
title: "Data Manipulation and Cleaning with Tidyverse"
format:
  revealjs:
    self-contained: true
    slide-number: true
editor: visual
---
<style>
.reveal {
  font-size: 160%;
}
</style>

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(stringr)
```

## Tidyverse

- **Tidyverse** is a collection of R packages designed for data science.
- It provides a consistent set of functions and syntax for data manipulation and visualization.
- Key packages include `dplyr`, `ggplot2`, `tidyr`, and `readr`.
  - `dplyr`: data manipulation
  - `ggplot2`: data visualization
  - `tidyr`: data cleaning
  - `readr`: data import
- Use `install.packages("tidyverse")` to install the package.
- Use `library(tidyverse)` to load the packages.

## Tidyverse `reader` package

- The `readr` package provides functions for reading data into R.
- The functions are similar to base R functions like `read.csv()`, but faster and more user-friendly.
- Common functions:
  - `read_csv()`: reads comma-separated files.
  - `read_tsv()`: reads tab-separated files.
  - `read_delim()`: reads files with custom delimiters.
  - `read_fwf()`: reads fixed-width files.
  - `read_table()`: reads tabular data.
- Returns a `tibble`, a modern version of a data frame.

## Sample data

```{r}
#| echo: true
#| include: true
library(tidyverse)

df <- tibble(
  `Gene Description` = c("Protein kinase C (alpha)",
                         "Ribosomal protein L2",
                         "ATP synthase", "Control gene"),
  `Gene Accession Number` = c("Gene1", "Gene2", "Gene3", "AFFX-Control"),
  Sample1 = c(5.2, 3.4, 7.8, 1.2),
  call1 = c("P", "A", "P", "A"),
  Sample2 = c(5.5, 3.8, 8.0, 1.3),
  call2 = c("P", "A", "P", "A")
)
```

## Sample data (cont'd)

```{r}
#| echo: true
#| include: true
print(dim(df))
print(colnames(df))
print(head(df))
```

## Using the Pipe Operator: `%>%`

- The `%>%` operator passes the result of one function into the next.

```{r}
#| echo: true
#| include: true
s <- rnorm(100)
m <- mean(s)

# Using the pipe operator
m <- s %>% mean()
# or
m <- rnorm(100) %>% mean()
# is equivalent to
m <- mean(rnorm(100))
```

## Using the Pipe Operator (%>%)

- Here, we select the "Gene Description" column and then display its head.

```{r}
#| echo: true
#| include: true
df %>%
  select(`Gene Description`) %>%
  head()

# This is equivalent to
head(select(df, `Gene Description`))
```

## `select()` function (dplyr package)

- The `select()` function is used to select columns from a data frame.
- Syntax: `select(data, column1, column2, ...)`
- Use `-` to exclude columns.
- Use `everything()` to include all remaining columns.
- Use `starts_with()`, `ends_with()`, `contains()`, `matches()` to select by name patterns.
- See `?select` for more details.

## `select()` function examples

```{r}
#| echo: true
#| include: true
# Select columns by name
df %>%
  select(`Gene Description`, `Gene Accession Number`) %>%
  head()
```
## `select()` function examples
```{r}
#| echo: true
#| include: true
# Remove columns using `-`
df %>%
  select(-`Gene Description`, -`Gene Accession Number`) %>%
  head()

# Remove columns using `-` and contains()
df %>%
  select(-contains("call")) %>%
  head()
```

> 📝 **Task 1** — remove columns whose names start with `"call"` using `starts_with()`

## Filtering rows with `filter()` function

- `filter()` is used to select rows based on conditions.
- Syntax: `filter(data, condition)`
- Use logical operators: `==`, `!=`, `>`, `<`, `>=`, `<=`, `%in%`, `&`, `|`, `!`
- Use `between()` to filter rows within a range.

```{r}
#| echo: true
#| include: true
kinase_df <- df %>%
  filter(str_detect(`Gene Description`, "kinase"))
print(kinase_df)
```

> 📝 **Task 2** — filter `df` to keep only rows where `Sample1 > 4`

## Creating new columns with `mutate()` function

- `mutate()` is used to create new columns or modify existing columns.
- Syntax: `mutate(data, new_column = expression)`

```{r}
#| echo: true
#| include: true
df_extended <- df %>%
  mutate(IsControl = str_detect(`Gene Accession Number`, "^AFFX"))
print(df_extended)
```

## Using `mutate()` to create new columns

- Calculate the average expression across Sample1 and Sample2.

```{r}
#| echo: true
#| include: true
df_extended <- df_extended %>%
  mutate(AvgExpr = rowMeans(select(., Sample1, Sample2), na.rm = TRUE))
print(df_extended)
```

> 📝 **Task 3** — add a column `"HighExpr"` that is `TRUE` when `AvgExpr > 5`

## Grouping and summarizing with `group_by()` and `summarize()`

- `group_by()` groups data by one or more variables.
- `summarize()` calculates summary statistics for each group.
- Use `ungroup()` to remove grouping.

```{r}
#| echo: true
#| include: true
by_cyl <- mtcars %>% group_by(cyl)
# Grouping does not change the data frame (it just adds a grouping attribute)
print(by_cyl)
```

## Grouping changes the behavior of functions

```{r}
#| echo: true
#| include: true
by_cyl <- mtcars %>% group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))
print(by_cyl)
by_cyl_gear <- mtcars %>% group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg))
print(by_cyl_gear)
```

## Grouping (cont'd)

```{r}
#| echo: true
#| include: true
control_summary <- df_extended %>%
  group_by(IsControl) %>%
  summarize(mean_expr = mean(AvgExpr), sd_expr = sd(AvgExpr))
print(control_summary)
```

> 📝 **Task 4** — calculate median horsepower (`hp`) per cylinder count in `mtcars`

## Reshaping data with `pivot_longer()` and `pivot_wider()` functions

- `pivot_longer()`: converts wide data to long data.
- `pivot_wider()`: converts long data to wide data.
- Wide data: multiple columns for each variable.
- Long data: multiple rows for each variable.

## Wide format example

```{r}
#| echo: true
#| include: true
wide_df <- tibble(
  ID = c(1, 2, 3, 4),
  Gene = c("Gene1", "Gene2", "Gene3", "Gene4"),
  Expr1 = c(5.2, 3.4, 7.8, 1.2),
  Expr2 = c(5.5, 3.8, 8.0, 1.3),
  Expr3 = c(5.5, 3.8, 8.0, 1.3),
  Expr4 = c(5.5, 3.8, 8.0, 1.3),
  Expr5 = c(5.5, 3.8, 8.0, 1.3)
)
print(wide_df)
```

## Long format example

- `pivot_longer(data, cols, names_to, values_to)`
  - `cols`: columns to pivot.
  - `names_to`: new column that stores original column names.
  - `values_to`: new column that stores the values.

```{r}
#| echo: true
#| include: true
long_df <- wide_df %>%
  pivot_longer(cols = starts_with("Expr"),
               names_to = "Sample",
               values_to = "Expression")
print(long_df)
```

## Converting long data to wide data

- `pivot_wider(data, names_from, values_from)`
  - `names_from`: column whose values become new column names.
  - `values_from`: column whose values fill the new columns.

```{r}
#| echo: true
#| include: true
wide_df <- long_df %>%
  pivot_wider(names_from = "Sample",
              values_from = "Expression")
print(wide_df)
```

> 📝 **Task 5** — add a `log2Expr` column to `long_df`, then pivot back to wide format

## Data joining with `left_join()`, `right_join()`, `inner_join()`, and `full_join()`

- `left_join()`: all rows from left + matched rows from right.
- `right_join()`: all rows from right + matched rows from left.
- `inner_join()`: only rows with matching values in both tables.
- `full_join()`: all rows from either table.

```{r}
#| echo: true
#| include: true
df1 <- tibble(ID = c(1, 2, 3), Value = c(10, 20, 30))
df2 <- tibble(ID = c(2, 3, 4), Value = c(200, 300, 400))
joined_df <- df1 %>% left_join(df2, by = "ID")
print(joined_df)
joined_df <- df1 %>% right_join(df2, by = "ID")
print(joined_df)
```

## Data joining (cont'd)

```{r}
#| echo: true
#| include: true
df1 <- tibble(ID = c(1, 2, 3), Value = c(10, 20, 30))
df2 <- tibble(ID = c(2, 3, 4), Value = c(200, 300, 400))

joined_df <- df1 %>% inner_join(df2, by = "ID")
print(joined_df)

joined_df <- df1 %>% full_join(df2, by = "ID")
print(joined_df)
```

> 📝 **Task 6** — create `df3 <- tibble(ID = c(1,3,5), Label = c("A","B","C"))` and `full_join()` it with `df1`. Which rows get `NA`? Why?

## String manipulation with functions from `stringr` package

- Common functions:
  - `str_detect()`: detect patterns in strings.
  - `str_replace()`: replace patterns in strings.
  - `str_split()`: split strings based on patterns.
  - `str_sub()`: extract substrings.
  - `str_trim()`: remove leading/trailing whitespace.
  - `str_to_lower()`, `str_to_upper()`: change case.
  - `str_c()`: concatenate strings.
  - `str_remove()`, `str_remove_all()`: remove patterns.

## `stringr` package examples

```{r}
#| echo: true
#| include: true
library(stringr)
str_detect("Hello, world!", "world")
str_replace("Hello, world!", "world", "universe")
str_split("Hello, world!", ",")
str_sub("Hello, world!", 3, 7)
str_trim("  Hello, world!  ")
str_remove("Hello, world!", "world")
```

> 📝 **Task 7** — strip `"_human"` / `"_mouse"` suffixes from `c("TP53_human", "BRCA1_human", "EGFR_mouse")` using `str_remove()`

## Regular expressions : defining search patterns

- Basic regex syntax:
  - `.`: any character; `^`: start; `$`: end
  - `[]`: character class; `|`: alternation
  - `*`: 0 or more; `+`: 1 or more; `?`: 0 or 1
  - `{n}`, `{n,}`, `{n,m}`: repetition counts
  - `\\`: escape; `\\d`: digit; `\\w`: word character

## Regular expressions examples

```{r}
#| echo: true
#| include: true
# 1. Dot (.)
str_detect("cat", "c.t")       # TRUE
str_detect("ct", "c.t")        # FALSE

# 2. Start (^) and End ($)
str_detect("Hello world", "^Hello")  # TRUE
str_detect("Hello world", "world$")  # TRUE

# 3. Character Class ([])
str_extract("hello", "[aeiou]")
str_extract_all("hello", "[aeiou]")
```

## Regular expressions examples (cont'd)

```{r}
#| echo: true
#| include: true
# 4. Alternation (|)
str_detect("I like cats", "cat|dog")    # TRUE
str_detect("I like dogs", "cat|dog")    # TRUE
str_detect("I like birds", "cat|dog")   # FALSE

# 5. Asterisk (*)
str_detect("ac", "ab*c")    # TRUE
str_detect("abbc", "ab*c")  # TRUE

# 6. Plus (+)
str_detect("abc", "ab+c")   # TRUE
str_detect("abbc", "ab+c")  # TRUE
str_detect("ac", "ab+c")    # FALSE
```

## Regular expressions examples (cont'd)

```{r}
#| echo: true
#| include: true
# 7. Question Mark (?)
str_detect("color", "colou?r")    # TRUE
str_detect("colour", "colou?r")   # TRUE

# 8. Exact Count ({n})
str_detect("aaa", "a{3}")   # TRUE
str_detect("aa", "a{3}")    # FALSE

# 9. Minimum Count ({n,})
str_detect("aa", "a{2,}")    # TRUE
str_detect("a", "a{2,}")     # FALSE
```

## Regular expressions examples (cont'd)

```{r}
#| echo: true
#| include: true
# 10. Range Count ({n,m})
str_detect("aa", "a{2,4}")    # TRUE
str_detect("aaaa", "a{2,4}")  # TRUE
str_detect("aaaaa", "a{2,4}") # FALSE

# 11. Escape Character (\\)
str_detect("a.b", "a\\.b")   # TRUE
str_detect("aXb", "a\\.b")   # FALSE
```

## Regular expressions examples (cont'd)

```{r}
#| echo: true
#| include: true
# 12. Digit (\\d)
str_extract("Order number: 12345", "\\d+")

# 13. Word Character (\\w)
str_extract_all("Hello, world!", "\\w+")
```

## Testing regular expressions

- Use online tools like [regex101](https://regex101.com/) to test regular expressions.

> 📝 **Task 8** — use `str_detect()` to extract only AFFX control probes from
> `c("AFFX-BioB-3_at", "Gene_1234", "AFFX-BioC-5_at", "Kinase_99")`