---
title: "Mutation heatmap — concepts you will use"
subtitle: "Reshaping a long mutation table, then aligning it to a tree with `gheatmap`"
author: "Instructor notes"
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 9
    fig-height: 7
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Render from the project root with:
#   quarto render notes/concepts.qmd
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r packages}
library(ape)
library(treeio)
library(ggtree)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(tibble)
```

# Why this document exists

This is a self-contained gallery of the techniques you need for the **mutation-heatmap** starter project. Every concept is illustrated with a plot generated from the actual project data. Read it once before you start so you know what the destination looks like; come back to it when `gheatmap` does something unexpected (it will — there is one specific trap that catches every first-time user).

The single question of this starter is: *did key spike mutations evolve convergently across SARS-CoV-2 lineages?* The visual answer is a heatmap aligned to a phylogenetic tree, painted by `ggtree::gheatmap`.

# 1. The tree as a 2-D coordinate system (recap)

A drawn tree has two axes:

- **y-axis** — indexes tips. Each tip has a unique y-coordinate, internal nodes sit at the average of their children's.
- **x-axis** — substitutions per site or calendar time. For SARS-CoV-2 builds, calendar time.

Once you accept this, **`gheatmap` is just adding extra columns of cells to the right of the tree**, where each cell row sits at a tip's y-coordinate and each cell column is one variable in your matrix. That mental model is enough to use it.

# 2. Loading the data

```{r load_data}
tree       <- read.tree("data/ncov_subsample.nwk")
meta       <- read_tsv("data/ncov_metadata.tsv", show_col_types = FALSE)
spike_long <- read_csv("data/spike_mutations_long.csv", show_col_types = FALSE)

cat("Tips in the tree:        ", length(tree$tip.label), "\n")
cat("Metadata rows:           ", nrow(meta), "\n")
cat("Long mutation table rows:", nrow(spike_long), "\n")
cat("Distinct spike mutations:", length(unique(spike_long$mutation)), "\n")
```

**Always check the join-key invariant first.** If even one strain name is mangled, every figure that follows is silently wrong.

```{r join_check}
length(setdiff(tree$tip.label, meta$strain))   # MUST be 0
length(setdiff(tree$tip.label, spike_long$strain))  # OK to be > 0:
                                                    # tips with no spike mutations
                                                    # are missing from the long table
```

The second number being non-zero is *expected* — a tip with zero observed spike mutations does not appear in the long table. We will fix that later by row-padding the wide matrix.

# 3. The long-format mutation table — what's in it

`data/spike_mutations_long.csv` is **long format**: one row per `(strain, mutation)` pair, restricted to spike (`S:`) substitutions. Each row says "this tip has this spike mutation."

```{r long_head}
spike_long %>% slice_sample(n = 6)
```

Reversions (a substitution and its reverse appearing on the same lineage) have already been collapsed for you, so a row in this table really does mean "the tip *carries* this mutation," not just "this mutation appeared somewhere along its lineage."

The most common spike substitutions in this 400-tip sample:

```{r mutation_counts}
spike_long %>%
  count(mutation, sort = TRUE) %>%
  head(15) %>%
  knitr::kable()
```

`S:D614G` is in nearly every tip — it fixed globally in early 2020 and never went away. The rest of the top of the list is dominated by Omicron-defining substitutions (`H655Y`, `K417N`, `S477N`, `Q498R`, …).

# 4. Long → wide: `pivot_wider`

`gheatmap` wants a **wide** matrix: rows are tips, columns are mutations, cell values are presence/absence. We start by picking the mutations we care about:

```{r key_muts}
key_muts <- c("S:D614G", "S:N501Y", "S:E484K", "S:E484A", "S:L452R",
              "S:T478K", "S:K417N", "S:K417T", "S:P681H", "S:P681R",
              "S:H655Y")
```

Then `pivot_wider`:

```{r pivot}
mut_wide <- spike_long %>%
  filter(mutation %in% key_muts) %>%
  mutate(present = "present") %>%
  pivot_wider(names_from = mutation,
              values_from = present,
              values_fill = "absent") %>%
  column_to_rownames("strain")

dim(mut_wide)
mut_wide[1:5, 1:5]
```

Two things to do *immediately* after the pivot, both essential for `gheatmap` to work:

## 4a. Force every `key_mut` to be a column

If a `key_mut` happens to be absent from every tip in the sample, `pivot_wider` silently drops the column. Force every key mutation to appear:

```{r ensure_columns}
for (m in setdiff(key_muts, colnames(mut_wide))) mut_wide[[m]] <- "absent"
mut_wide <- mut_wide[, key_muts]   # also enforces a stable column order
colnames(mut_wide)
```

## 4b. Force every tip to be a row

Tips with **zero** of our `key_muts` were dropped by the pivot (they have no rows in the long table after the `filter`). They still have to appear in the matrix or `gheatmap` will misalign:

```{r pad_rows}
missing_tips <- setdiff(tree$tip.label, rownames(mut_wide))
length(missing_tips)

if (length(missing_tips) > 0) {
  pad <- as.data.frame(
    matrix("absent",
           nrow = length(missing_tips),
           ncol = ncol(mut_wide),
           dimnames = list(missing_tips, colnames(mut_wide)))
  )
  mut_wide <- rbind(mut_wide, pad)
}

cat("Wide matrix:", nrow(mut_wide), "rows ×", ncol(mut_wide), "columns\n")
cat("All tips present as rows?",
    setequal(rownames(mut_wide), tree$tip.label), "\n")
```

# 5. The base tree (foundation for the heatmap)

Build a time-calibrated `ggtree` with tip color by clade. The tip color is *not* the answer to the question — its job is to give the heatmap rows clade context so you can read which lineage a row belongs to.

```{r base_tree}
meta <- meta %>% mutate(decimal_date = decimal_date(ymd(date)))

td <- full_join(as_tibble(tree), meta, by = c("label" = "strain")) %>%
  as.treedata()

mrsd_value <- max(meta$decimal_date, na.rm = TRUE)

p1 <- ggtree(td, mrsd = mrsd_value, as.Date = TRUE,
             size = 0.25, color = "grey40") +
  geom_tippoint(aes(color = Nextstrain_clade), size = 1.1, alpha = 0.85) +
  scale_color_viridis_d(option = "turbo", name = "Clade",
                        guide = guide_legend(ncol = 2,
                                             override.aes = list(size = 3))) +
  theme_tree2() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1
```

**Branch-units sanity check** — `mrsd` only works if branches are in years:

```{r units}
max(node.depth.edgelength(tree))   # should be ~6 here
```

# 6. `gheatmap` — the answer figure

`gheatmap` glues the wide matrix to the right of `p1`. Each cell row sits at a tip's y-coordinate, each column is one mutation, each cell color is `present` or `absent`.

```{r gheatmap, fig.height=9}
p_heat <- gheatmap(p1, mut_wide,
                   offset = 50,            # gap between tree and heatmap
                   width  = 0.6,           # heatmap width relative to tree
                   colnames_angle = 45,
                   colnames_offset_y = -2,
                   font.size = 3) +
  scale_fill_manual(values = c(present = "firebrick", absent = "grey92"),
                    name = "Spike mutation") +
  ggtitle("Convergent vs clonal patterns of spike substitutions")

p_heat
```

# 7. Reading the heatmap — clonal vs convergent

This is what the figure is *for*. Look down each mutation column:

- **One coherent block of red cells** = the mutation arose once, on a single ancestral branch, and was inherited by every descendant. This is **clonal descent** — boring, but the null hypothesis.
- **Multiple separated red blocks** at different parts of the tree = the mutation arose **independently** on multiple branches. This is **convergent evolution** — the visual fingerprint of selection acting in parallel.

Examples in this dataset:

```{r convergence_summary, echo=FALSE}
n_clades_with <- function(mut) {
  if (!mut %in% colnames(mut_wide)) return(0)
  tips_with <- rownames(mut_wide)[mut_wide[[mut]] == "present"]
  length(unique(meta$Nextstrain_clade[meta$strain %in% tips_with]))
}

tibble(
  Mutation = key_muts,
  `Tips carrying it` = vapply(key_muts, function(m) sum(mut_wide[[m]] == "present"), integer(1)),
  `Distinct clades carrying it` = vapply(key_muts, n_clades_with, integer(1))
) %>% knitr::kable()
```

A high value in the third column is a quantitative cousin of "multiple separated blocks" — the mutation has been picked up across many independent clades. Compare a few:

- **`S:D614G`** is in *every* clade after 19A/19B — single block at the top of the tree, then fixed. This is the **near-universal** post-2020 substitution. Visually: one giant red block.
- **`S:N501Y`** appears in **Alpha (`20I`)**, in **Beta (`20H`)**, and in much of **Omicron (`21K`, `21L`, `22B`, …)**. Visually: multiple separated blocks. This is the textbook convergence pattern.
- **`S:L452R`** appears in Delta (`21A`/`21I`/`21J`) and again in some Omicron sublineages (e.g. BA.5 / `22B`) — also multiple blocks.

When you write your interpretation, name *specific* mutations and the *specific* lineages where they recur.

# 8. Why row-name alignment is the trap that bites everyone

`gheatmap` aligns the matrix to the tree by **row name**, not row order. Three ways to break this:

1. **Sorting the matrix manually** (e.g. `mut_wide %>% arrange(strain)`). This re-orders rows but `gheatmap` will *still* use row names — so visually nothing changes — but if you accidentally lose row names during the sort (`%>% arrange(rownames(.))` is a recipe for that), the matrix silently misaligns.
2. **Forgetting to pad missing tips.** Tips with no key mutations are absent from the pivoted matrix → those tips get *no row* in the heatmap and downstream rows shift up one position to fill the gap. The figure looks fine but is wrong.
3. **Forgetting to force every column.** A `key_mut` absent from the sample → the column doesn't exist → either `gheatmap` errors, or the column ordering shifts and your column labels lie.

The only safe pattern is the one in §4: force columns with the `setdiff` loop, force rows with the `rbind` of an `"absent"` data frame.

# 9. Pitfalls quick-reference

```{r pitfalls, echo=FALSE, results='asis'}
tab <- tribble(
  ~Symptom, ~Cause, ~Fix,
  "Heatmap rows misaligned with tips",
    "`gheatmap` uses row *names*, not row order; or some tips are missing rows",
    "Don't sort manually; pad missing tips with all-`absent` rows",
  "A `key_mut` you expected is missing from the heatmap",
    "It's absent from every tip in this sample → `pivot_wider` dropped the column",
    "Force the column: `for (m in setdiff(key_muts, colnames(mut_wide))) mut_wide[[m]] <- 'absent'`",
  "X-axis on the tree is wrong / labels pre-1900",
    "Branches in subs/site, not time",
    "`max(node.depth.edgelength(tree))` must be a few years; otherwise time-calibrate",
  "Strain join silently drops half the tips",
    "Strain-name encoding mismatch (`/`, `|`, whitespace)",
    "Run `setdiff(tree$tip.label, meta$strain)` before joining",
  "Heatmap looks fine but `gheatmap` warns about Date conversion",
    "Harmless; ggtree converts the numeric x to Date for the time axis",
    "Ignore"
)
knitr::kable(tab)
```

# 10. Where to go next

When this project is done, the natural next steps live in sibling directories:

- **`SARS-CoV_clade_waves/`** — same tree, different question: *did clades replace each other through time?* No `gheatmap`; the answer is in tip color alone.
- **`SARS-CoV_analysis_project/`** — the full version, where `gheatmap` is one of six visual encodings layered on the same tree (tip color, branch color, heatmap, facet strips, MSA, geographic projection).

# Session info

```{r sessioninfo, echo=FALSE}
sessionInfo()
```
