---
title: "Biostrings, BSgenome, and Bioconductor"
format:
  revealjs:
    self-contained: true
    slide-number: true
editor: visual
---
<style>
.reveal {
  font-size: 160%;
}
</style>

```{r}
#| echo: false
#| include: false
library(Biostrings)
library(BSgenome)
library(GenomicRanges)

# Compatibility shim: pairwiseAlignment + BLOSUM/PAM were moved to the
# pwalign package in Bioc 3.20+. Load pwalign if present; otherwise fall
# back to Biostrings.
if (requireNamespace("pwalign", quietly = TRUE)) {
  library(pwalign)
}
load_sub_matrix <- function(name) {
  if (requireNamespace("pwalign", quietly = TRUE)) {
    get(name, envir = asNamespace("pwalign"))
  } else {
    env <- new.env()
    data(list = name, package = "Biostrings", envir = env)
    env[[name]]
  }
}
```

# Part 0 — Installation

## Installing Biostrings, BSgenome, pwalign

All packages come from Bioconductor via `BiocManager`:

```{r}
#| eval: false
#| echo: true
#| include: true
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "Biostrings",
  "BSgenome",
  "GenomicRanges",
  "pwalign",                         # Bioc 3.20+ only (see next slide)
  "BSgenome.Hsapiens.UCSC.hg38",     # optional — large (~1 GB)
  "BSgenome.Drerio.UCSC.danRer11"    # optional
))
```

```{r}
#| echo: true
#| include: true
BiocManager::version()
```

## Heads-up — the `pwalign` split

In **Bioconductor 3.20 (Oct 2024)** the `pairwiseAlignment()` function and the
substitution-matrix data (`BLOSUM*`, `PAM*`) were moved out of Biostrings
into a new package called **`pwalign`**.

-   **Bioc ≤ 3.19** (this teaching host is 3.18): they still live in Biostrings — no extra `library()` needed.
-   **Bioc ≥ 3.20** (classroom workstations are 3.22): install `pwalign` and `library(pwalign)` *before* calling `pairwiseAlignment()` / using `BLOSUM62`.

```{r}
#| eval: false
#| echo: true
#| include: true
# Same code works on both versions if you guard the pwalign load:
if (requireNamespace("pwalign", quietly = TRUE)) {
  library(pwalign)           # Bioc >= 3.20
}
# On Bioc <= 3.19 nothing extra is needed.
```

---

# Part 1 — Biostrings

## Biostrings: Sequence Analysis in R

-   High-performance classes for biological sequences:
    -   `DNAString`, `RNAString`, `AAString`
    -   `DNAStringSet`, `RNAStringSet`, `AAStringSet` (collections)
-   Fast pattern matching, alignment, and frequency operations.
-   Integrates tightly with `GenomicRanges` and `BSgenome`.

## Creating a Sequence Object

```{r}
#| echo: true
#| include: true
dna_seq <- DNAString("ATGCGTACGTAGCTAG")
print(dna_seq)

rna_seq <- RNAString("AUGCGUACGUAGCUAG")
print(rna_seq)

rna_converted <- as(rna_seq, "RNAString")
print(rna_converted)

aa_seq <- AAString("MVLSPADKTNVKAAW")
print(aa_seq)

dna_set <- DNAStringSet(c("ATGCGTACGACAGTAGCTAG",
                          "GATTACAAACATAAA",
                          "TTACATGACCCTTTACATG"))
print(dna_set)
```

> 📝 **Task 1** — build a named `DNAStringSet` with three sequences
> (`geneA`, `geneB`, `geneC`); print the object and its `names()`.

## Basic Sequence Operations

-   `width()` — length of each sequence (or `nchar()` on character strings)
-   `subseq()` — extract a subsequence
-   `reverseComplement()` — reverse complement a DNA sequence

```{r}
#| echo: true
#| include: true
seq_length <- width(dna_set)
print(seq_length)

subseq_dna <- subseq(dna_set, start = 4, end = 8)
print(subseq_dna)
```

## Reverse Complement

```{r}
#| echo: true
#| include: true
rev_comp_dna <- reverseComplement(dna_set)
print(dna_set)
print(rev_comp_dna)
```

## Translation

`translate()` converts a nucleotide sequence to its protein sequence.

```{r}
#| eval: false
#| echo: true
#| include: true
translate(x, genetic.code = GENETIC_CODE,
          no.init.codon = FALSE, if.fuzzy.codon = "error")
```

`GENETIC_CODE` is the standard code; pass a different table for mitochondrial
or alternative codes.

```{r}
#| echo: true
#| include: true
head(GENETIC_CODE, 16)
translated_seq <- translate(dna_set)
print(translated_seq)
```

> 📝 **Task 2** — take the reverse complement of `dna_set` and translate it.

## Pattern Matching — Functions

```{r}
#| eval: false
#| echo: true
#| include: true
# single sequence
matchPattern(pattern, subject,
             max.mismatch = 0, min.mismatch = 0,
             with.indels = FALSE, fixed = TRUE, algorithm = "auto")
countPattern(pattern, subject,
             max.mismatch = 0, min.mismatch = 0,
             with.indels = FALSE, fixed = TRUE, algorithm = "auto")

# multiple sequences (vectorized)
vmatchPattern(pattern, subject,
              max.mismatch = 0, min.mismatch = 0,
              with.indels = FALSE, fixed = TRUE, algorithm = "auto")
vcountPattern(pattern, subject,
              max.mismatch = 0, min.mismatch = 0,
              with.indels = FALSE, fixed = TRUE, algorithm = "auto")
```

## Pattern Matching in a Single Sequence

```{r}
#| echo: true
#| include: true
pattern <- "CGTACG"
print(dna_seq)

match_result <- matchPattern(pattern, dna_seq)
print(match_result)

pattern_count <- countPattern(pattern, dna_seq)
print(paste("Number of occurrences of", pattern, ":", pattern_count))
```

## Pattern Matching in a Set of Sequences

```{r}
#| echo: true
#| include: true
pattern <- "ACAT"
print(dna_set)

match_result_set <- vmatchPattern(pattern, dna_set)
print(match_result_set)
```

## Matching with Mismatches

```{r}
#| echo: true
#| include: true
pattern <- "ACAT"
dna_seq_mm <- DNAString("ACATGACATGACATGACGT")
match_result_mm <- matchPattern(pattern, dna_seq_mm, max.mismatch = 1)
print(match_result_mm)
```

> 📝 **Task 3** — search `dna_set` for `"GTAG"` allowing up to one mismatch;
> how many matches are found per sequence? Hint: `vcountPattern()`.

## Reading FASTA Files

-   `readDNAStringSet()` / `readAAStringSet()` read sequences from FASTA.
-   `writeXStringSet()` writes them back out.
-   Plain text and gzipped files are handled transparently.

```{r}
#| echo: true
#| include: true
fasta_file_1 <- "../../data/seq_data/someORF.fa"
fasta_file_2 <- "../../data/seq_data/dm3_upstream2000.fa.gz"

dna_seq_fasta <- readDNAStringSet(fasta_file_1)
print(dna_seq_fasta)

dna_seq_fasta_gz <- readDNAStringSet(fasta_file_2)
print(dna_seq_fasta_gz)
```

## Reading Protein FASTA

```{r}
#| echo: true
#| include: true
protein_fasta_file <- "../../data/seq_data/UP000000625.fasta.gz"
protein_seq_fasta <- readAAStringSet(protein_fasta_file)
print(protein_seq_fasta)
```

## Importing Multiple Sequence Alignments

-   `readDNAMultipleAlignment()` / `readAAMultipleAlignment()`
-   Supported formats: `"fasta"`, `"clustal"`, `"phylip"`

```{r}
#| echo: true
#| include: true
aln_file <- "../../data/seq_data/example_msa.fasta"
aln <- readAAMultipleAlignment(aln_file, format = "fasta")
print(aln)
```

## Consensus from an MSA

-   `consensusMatrix()` — per-position letter counts
-   `consensusString()` — single consensus sequence

```{r}
#| echo: true
#| include: true
cons_mat <- consensusMatrix(aln)
print(cons_mat[, 1:10])

consensus_seq <- consensusString(cons_mat)
print(consensus_seq)
```

> 📝 **Task 4** — compute `consensusMatrix(aln, as.prob = TRUE)` to get
> per-position letter *frequencies* rather than counts.

## Pairwise Alignment

-   `pairwiseAlignment()` aligns two sequences.
-   Type: `"global"` (default), `"local"`, or `"overlap"`.

```{r}
#| echo: true
#| include: true
seq1 <- DNAString("ATGCTAGCTAG")
seq2 <- DNAString("ATGCGGCTAG")

alignment <- pairwiseAlignment(seq1, seq2)
print(alignment)

aligned_seq1 <- pattern(alignment)
aligned_seq2 <- subject(alignment)
print(aligned_seq1)
print(aligned_seq2)
print(paste("Alignment score:", score(alignment)))
```

> 📝 **Task 5** — run the same pair with `type = "local"` and compare scores.

## Scoring — substitution matrices for proteins

The default score penalises every mismatch equally. For proteins that is
biologically wrong: a conservative substitution (I↔L, K↔R) should cost less
than a disruptive one (W↔D).

**BLOSUM** and **PAM** matrices encode the evolutionary probability of each
amino-acid substitution, built from databases of real homologous proteins.

```{r}
#| echo: true
#| include: true
BLOSUM62 <- load_sub_matrix("BLOSUM62")
BLOSUM45 <- load_sub_matrix("BLOSUM45")

BLOSUM62[1:6, 1:6]
```

## Same pair, different scoring model

Two short fragments around the heme pocket of alpha- and beta-globin:

```{r}
#| echo: true
#| include: true
p1 <- AAString("PADAVMGNPKVKAHGKKVLGA")
p2 <- AAString("PKVKAHGKKVLGAFSDGLAHL")

a62 <- pairwiseAlignment(p1, p2, substitutionMatrix = BLOSUM62,
                         gapOpening = 10, gapExtension = 4)
a45 <- pairwiseAlignment(p1, p2, substitutionMatrix = BLOSUM45,
                         gapOpening = 10, gapExtension = 4)

cat("BLOSUM62 score:", score(a62), "\n")
cat("BLOSUM45 score:", score(a45), "\n")
```

BLOSUM45 is calibrated for **more distant** homologs, and gives a higher
score here. **The choice of matrix is part of the biological question.**

## Finding palindromes — `findPalindromes`

A palindrome in DNA = a region that equals its **reverse complement**.
Two biological flavours:

-   **Restriction sites** are palindromic by design (EcoRI `GAATTC`, BamHI `GGATCC`, HindIII `AAGCTT`, …).
-   **Hairpins / cruciforms** = two palindromic arms with a small loop in the middle — common in RNA secondary structure.

## Restriction-site scan (strict palindrome)

```{r}
#| echo: true
#| include: true
plasmid <- DNAString("AAAGAATTCCCCGAATTCAAGGATCCCCC")

res_sites <- findPalindromes(plasmid,
                             min.armlength = 3,
                             max.looplength = 0,
                             max.mismatch = 0)
res_sites
```

Two EcoRI (`GAATTC`) and one BamHI (`GGATCC`) site — found in one call.

## Hairpin-style palindromes (small loop allowed)

```{r}
#| echo: true
#| include: true
rna_like <- DNAString("GGGAAATTTCCCAAAGGGTTTCCC")

hairs <- findPalindromes(rna_like,
                         min.armlength = 3,
                         max.looplength = 5,
                         max.mismatch = 0)
hairs
palindromeArmLength(hairs)
```

Helpers: `palindromeArmLength()`, `palindromeLeftArm()`, `palindromeRightArm()`.

## Indexing XStringSet

-   `[` returns an `XStringSet` (a subset — still a collection).
-   `[[` returns an `XString` (a single unwrapped sequence).
-   Accessing by name also works.

```{r}
#| echo: true
#| include: true
first_seq <- dna_set[1]
print(first_seq)

first_seq_2 <- dna_set[[1]]
print(first_seq_2)

multiple_seqs <- dna_set[1:2]
print(multiple_seqs)
```

## Nucleotide Frequencies

-   `alphabetFrequency()` — counts for every letter of the alphabet
-   `letterFrequency()` — counts for specified letters (incl. `"GC"`)
-   `oligonucleotideFrequency()` — k-mer counts

```{r}
#| echo: true
#| include: true
print(dna_seq)

nucl_freq <- alphabetFrequency(dna_seq, baseOnly = TRUE)
print(nucl_freq)
```

## Top 4-mers

```{r}
#| echo: true
#| include: true
oligo_freq <- oligonucleotideFrequency(dna_seq, width = 4)
sorted_oligo_freq <- sort(oligo_freq, decreasing = TRUE)[1:10]
print(sorted_oligo_freq)
```

> 📝 **Task 6** — compute GC content of each sequence in `dna_set` as a
> proportion. Hint: `letterFrequency(dna_set, "GC", as.prob = TRUE)`.

---

# Part 2 — BSgenome

## Overview of BSgenome

-   Provides full **reference genomes** as Bioconductor data packages.
-   Chromosomes stored as `DNAString` objects, loaded **lazily from disk**.
-   Integrates directly with Biostrings (`getSeq`, `letterFrequency`, …) and
    GenomicRanges (subset by `GRanges` intervals).

## Installing BSgenome Packages

Install the infrastructure:

```{r}
#| echo: true
#| include: true
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("BSgenome")
```

Install specific genome packages:

```{r}
#| echo: true
#| include: true
#| eval: false
BiocManager::install(c(
  "BSgenome.Hsapiens.UCSC.hg38",
  "BSgenome.Drerio.UCSC.danRer11"
))
```

## Discovering Available Genomes

```{r}
#| echo: true
#| include: true
#| eval: false
available.genomes() |> head(10)   # all BSgenome packages in your release
installed.genomes()               # installed on this machine
```

## Inspecting a Genome

```{r}
#| echo: true
#| include: true
#| eval: false
library(BSgenome.Hsapiens.UCSC.hg38)
genome_human <- BSgenome.Hsapiens.UCSC.hg38

seqnames(genome_human) |> head(10)    # chromosome names
seqlengths(genome_human)["chr1"]      # chromosome length
```

## Basic Genome Queries

```{r}
#| echo: true
#| include: true
#| eval: false
# Extract a 10 kb window from chr1
chr1_10kb <- getSeq(genome_human, "chr1", 100000, 110000)
print(chr1_10kb)

# Access a whole chromosome
chr1 <- genome_human$chr1
print(chr1)

# Inspect provenance
metadata(genome_human)
```

## Extracting Subsequences with `getSeq()`

`getSeq()` also works on a plain `DNAStringSet` with chromosome-like names — useful for quick demos without installing a full BSgenome package.

```{r}
#| echo: true
#| include: true
g <- DNAStringSet(c(chr1 = "ACGTACGTACGT",
                    chr2 = "TGCACTGCA",
                    chr3 = "GATCGATC"))
print(g)

s1 <- getSeq(g, GRanges("chr1", IRanges(1, 4)))
print(s1)

s2 <- getSeq(g, GRanges("chr2", IRanges(start = 1:4, end = 4:6)))
print(s2)
```

## Benefits of BSgenome

-   **Reproducibility** — versioned, release-tied data packages
-   **Efficiency** — on-disk 2bit storage, lazy chromosome loading
-   **Discoverability** — `available.genomes()`, `installed.genomes()`,
    `getBSgenome()`
-   **Integration** — direct compatibility with Biostrings and GenomicRanges

## TxDb — Transcript Databases

A **TxDb** is an on-disk SQLite store of transcript annotations:

-   genes, transcripts, exons, CDS, UTRs

for a given genome build. Shipped as packages — e.g.:

-   `TxDb.Scerevisiae.UCSC.sacCer3.sgdGene`
-   `TxDb.Dmelanogaster.UCSC.dm6.ensGene`

Catalog: [bioconductor.org/packages/release/data/annotation/](https://bioconductor.org/packages/release/data/annotation/)

## Installing and Loading a TxDb

```{r}
#| echo: true
#| include: true
#| eval: false
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("TxDb.Dmelanogaster.UCSC.dm6.ensGene")

library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
txdb <- TxDb.Dmelanogaster.UCSC.dm6.ensGene
```

## Core TxDb Extraction Functions

-   `genes(txdb)` — GRanges of all genes
-   `transcripts(txdb)` — GRanges of all transcripts (with strand)
-   `exonsBy(txdb, by = "gene")` — GRangesList of exons per gene
-   `intronsByTranscript(txdb, use.names = TRUE)` — introns per transcript
-   `cdsBy(txdb, by = "tx")` — CDS per transcript
-   `promoters(txdb, upstream = 2000, downstream = 200)` — promoter windows

All return `GRanges` or `GRangesList`, which feed straight into `getSeq()` or
overlap-based functions in `GenomicRanges`.

---

# Part 3 — Bioconductor: the shared infrastructure

## What is Bioconductor?

-   Open-source project providing R packages for **high-throughput genomic data**.
-   Every package we used in this session — **Biostrings**, **BSgenome**, **GenomicRanges** — is part of Bioconductor.
-   They all build on the same foundation: **S4 classes, slots, and generic methods**.
-   Let's look at what that foundation actually is.

## Everything is an S4 object

The objects we've been creating are not plain lists or character vectors — they are **S4 objects** with defined classes.

```{r}
#| echo: true
#| include: true
class(dna_set)     # DNAStringSet
class(aln)         # AAMultipleAlignment
class(seq1)        # DNAString

isS4(dna_set)
isS4(aln)
```

## Slots — the internal storage

S4 objects store their data in named **slots**. Peek at them with `slotNames()` — but rarely touch them directly.

```{r}
#| echo: true
#| include: true
slotNames(dna_set)      # internal fields of a DNAStringSet
slotNames(aln)          # internal fields of an AAMultipleAlignment
```

Think of slots as "private fields" — the package decides the layout; you use **accessor functions** to read them.

## Accessors, not `$`

For S4 objects, `$` does **not** work the same way as for data frames. Use the package's accessor functions.

```{r}
#| echo: true
#| include: true
# Correct — use accessors
width(dna_set)
names(g)
alphabet(dna_seq)
```

```{r}
#| eval: false
#| echo: true
#| include: true
# Wrong — does not give "width" of the set
dna_set$width
```

Why? Accessors enforce the class contract — they keep working even if the internal slot layout changes.

## Methods dispatch on class

A generic function (`print`, `width`, `length`, `getSeq`, ...) runs a **different method** depending on the class of its argument.

```{r}
#| echo: true
#| include: true
head(methods(class = "DNAStringSet"), 10)
head(methods(class = "MultipleAlignment"), 10)
```

```{r}
#| echo: true
#| include: true
length(dna_set)    # number of sequences in the set
length(dna_seq)    # number of characters in the one sequence

width(dna_set)     # length of each sequence in the set
```

Same verb, different meaning per class. (`width()` is defined for the *set*, not for a single `DNAString`.)

## Generic functions across packages

One verb works across many classes, even across **different packages**. That is what lets Bioconductor compose:

```{r}
#| echo: true
#| include: true
# On a plain DNAStringSet with "chromosome" names:
getSeq(g, GRanges("chr1", IRanges(1, 4)))
```

```{r}
#| eval: false
#| echo: true
#| include: true
# Same call pattern on a BSgenome object:
getSeq(BSgenome.Hsapiens.UCSC.hg38, "chr1", 100000, 110000)

# Or on an indexed FASTA (FaFile):
getSeq(FaFile("genome.fa"), GRanges("chr1", IRanges(1, 1000)))
```

## Finding help and vignettes

How to keep learning after the course ends:

```{r}
#| eval: false
#| echo: true
#| include: true
?DNAStringSet               # help page for a class
methods("translate")        # every class translate() dispatches on
showMethods("getSeq")       # signatures of an S4 generic

vignette(package = "Biostrings")   # list a package's vignettes
browseVignettes("Biostrings")      # open them in a browser
```

**Vignettes** are worked-example tutorials — usually the fastest way to learn a new Bioconductor package.

## Summary

-   **Biostrings** — sequences: creation, manipulation, patterns, alignment, frequencies.
-   **BSgenome** — full reference genomes, reproducible and lazily loaded.
-   **TxDb** — transcript-level annotations, bridging sequences and gene models.
-   **Bioconductor** — the shared S4 infrastructure (classes, slots, generic methods) that makes all of these packages compose.

Further reading:

-   [Bioconductor](https://bioconductor.org/)
-   [Biostrings vignettes](https://bioconductor.org/packages/release/bioc/html/Biostrings.html)
-   [BSgenome vignettes](https://bioconductor.org/packages/release/bioc/html/BSgenome.html)
