---
title: "Example 2 — GC content around transcription start sites"
format:
  revealjs:
    self-contained: true
    slide-number: true
editor: visual
---
<style>
.reveal {
  font-size: 160%;
}
</style>

```{r setup}
#| echo: false
#| include: false
library(Biostrings)
library(BSgenome)
library(rtracklayer)
library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Drerio.UCSC.danRer11)
library(ggplot2)
```

## Installing the genome packages

Both BSgenome packages are ~500–900 MB and distributed via Bioconductor.
They are usually pre-installed on classroom workstations — if not, install
them once with `BiocManager`:

```{r}
#| eval: false
#| echo: true
#| include: true
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "BSgenome.Hsapiens.UCSC.hg38",     # human, ~900 MB
  "BSgenome.Drerio.UCSC.danRer11",   # zebrafish, ~450 MB
  "rtracklayer"                      # BED import
))
```

Check what's already installed:

```{r}
#| echo: true
#| include: true
requireNamespace("BSgenome.Hsapiens.UCSC.hg38", quietly = TRUE)
```

## Load libraries

```{r}
#| eval: false
#| echo: true
#| include: true
library(Biostrings)
library(BSgenome)
library(rtracklayer)
library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Drerio.UCSC.danRer11)
library(ggplot2)
```

## The biological question

-   Protein-coding genes in vertebrates are typically embedded in **GC-rich regions**, with the peak right at the transcription start site (TSS).
-   That GC peak overlaps with **CpG islands** and correlates with:
    -   open chromatin,
    -   efficient mRNA nuclear export,
    -   higher translation efficiency.
-   Plan for this example:
    1.  Extract ±1 kb windows around TSSs in **human** and **zebrafish**.
    2.  Compute GC content in a 50 bp sliding window.
    3.  Average across thousands of TSSs.
    4.  Compare against the chromosome-wide GC baseline.

## Loading genomes as BSgenome objects

Unlike Example 1, we use **precompiled BSgenome packages** here — no FASTA parsing, no re-downloading, chromosomes loaded lazily from disk.

```{r}
#| echo: true
#| include: true
BSgenome.Hsapiens.UCSC.hg38
```

## Peeking at the genome object

```{r}
#| echo: true
#| include: true
seqnames(BSgenome.Hsapiens.UCSC.hg38) |> head()
seqlengths(BSgenome.Hsapiens.UCSC.hg38)["chr1"]
```

-   A BSgenome behaves like a **named collection of chromosomes**.
-   Chromosome sequences are read from disk only when accessed — large
    genomes take very little RAM until you actually touch them.

## Importing BED files with rtracklayer

The TSS coordinates were pre-computed and saved as BED. Each row is a ±1 kb window centered on a TSS → width = 2001 bp.

```{r}
#| echo: true
#| include: true
hsapiens_bed_path <- "../../data/seq_data/tss_hs.bed"
drerio_bed_path   <- "../../data/seq_data/tss_drerio.bed"

hs_bed     <- import(hsapiens_bed_path, format = "BED")
drerio_bed <- import(drerio_bed_path,   format = "BED")

length(hs_bed)
width(hs_bed)[1:5]
hs_bed
```

## Aligning seqnames — the "chr" prefix problem

A very common pitfall:

-   BED / Ensembl style → `"1"`, `"2"`, `"X"`, `"Y"`
-   UCSC / BSgenome style → `"chr1"`, `"chr2"`, `"chrX"`, `"chrY"`

`getSeq()` fails or returns nothing when the names don't match **exactly**.

```{r}
#| echo: true
#| include: true
seqlevels(hs_bed) |> head()
seqlevels(BSgenome.Hsapiens.UCSC.hg38) |> head()
```

## Fixing the seqnames

```{r}
#| echo: true
#| include: true
seqlevels(hs_bed)     <- paste0("chr", seqlevels(hs_bed))
seqlevels(drerio_bed) <- paste0("chr", seqlevels(drerio_bed))

seqlevels(hs_bed) |> head()
```

Now both objects use the same naming convention — `getSeq()` will work.

## Extracting sequences with `getSeq()`

`getSeq(genome, ranges)` returns a `DNAStringSet`, one sequence per range.
Strand is respected — minus-strand ranges are **reverse-complemented** automatically.

```{r}
#| echo: true
#| include: true
hs_sequences     <- getSeq(BSgenome.Hsapiens.UCSC.hg38, hs_bed)
drerio_sequences <- getSeq(BSgenome.Drerio.UCSC.danRer11, drerio_bed)

hs_sequences
```

## `letterFrequencyInSlidingView()`

For a single sequence and a window of width `W`:

-   slides a window of width `W` across the sequence,
-   returns, for each window position, the count of the requested letters.

Output is a matrix with `length(seq) − W + 1` rows, one column per letter set.

```{r}
#| echo: true
#| include: true
n <- 5000
x <- sapply(sample(hs_sequences, n), function(seq) {
  letterFrequencyInSlidingView(seq, view.width = 50, letters = "GC")
})
dim(x)   # rows = window positions, cols = sampled TSSs
```

## Average GC profile around human TSS

```{r}
#| echo: true
#| include: true
#| fig-width: 9
#| fig-height: 5
gc_mean_content <- rowMeans(x, na.rm = TRUE)
plot(gc_mean_content,
     xlab = "Window start (bp)",
     ylab = "GC count in 50 bp window",
     main = "Mean GC content around human TSS")
n_windows <- length(gc_mean_content)
abline(v = n_windows / 2, col = "red", lty = 2)
```

The peak sits right at the TSS — exactly where we expected.

## A reusable profile function

The same logic works for any letter set (GC, AT, CG, single bases) and any window width. Wrap it up once, reuse it everywhere.

```{r}
#| echo: true
#| include: true
get_N_profile <- function(s, W = 50, letters = "GC", sample_size = 5000) {
  x <- sapply(sample(s, sample_size), function(seq) {
    letterFrequencyInSlidingView(seq, view.width = W, letters = letters)
  })
  mean_content <- rowMeans(x, na.rm = TRUE)
  mean_content / W          # normalise to a per-base proportion in [0, 1]
}

hs_gc_profile <- get_N_profile(hs_sequences,     W = 50, letters = "GC")
dr_gc_profile <- get_N_profile(drerio_sequences, W = 50, letters = "GC")
```

> 📝 **Task 1** — compute separate A, T, G, and C profiles for the human set
> and overlay them. Which bases drive the peak at the TSS?

## Comparing species with base R

```{r}
#| echo: true
#| include: true
#| fig-width: 9
#| fig-height: 5
plot(dr_gc_profile, type = "l", col = "blue", ylim = c(0, 1),
     xlab = "Window start (bp)", ylab = "GC content",
     main = "GC profile around TSS")
points(hs_gc_profile, type = "l", col = "red")
```

## Genome-wide GC baseline

Is the TSS peak really unusual? We need a reference — the chromosome-wide GC content, measured once on chr1 of each species.

```{r}
#| echo: true
#| include: true
gc_hs_chr1 <- letterFrequency(BSgenome.Hsapiens.UCSC.hg38$chr1,
                              letters = "GC", as.prob = TRUE)
gc_dr_chr1 <- letterFrequency(BSgenome.Drerio.UCSC.danRer11$chr1,
                              letters = "GC", as.prob = TRUE)

cat("Human   chr1 GC:", round(gc_hs_chr1, 3), "\n")
cat("Zebrafish chr1 GC:", round(gc_dr_chr1, 3), "\n")
```

Both TSS peaks sit well above each species' own genome-wide baseline.

## Baselines on the plot

```{r}
#| echo: true
#| include: true
#| fig-width: 9
#| fig-height: 5
plot(dr_gc_profile, type = "l", col = "blue", ylim = c(0, 1),
     xlab = "Window start (bp)", ylab = "GC content",
     main = "GC profile around TSS (with chr1 baselines)")
points(hs_gc_profile, type = "l", col = "red")
abline(h = gc_hs_chr1, col = "red",  lty = 2)
abline(h = gc_dr_chr1, col = "blue", lty = 2)
```

## The same figure with ggplot2 — code

```{r}
#| echo: true
#| include: true
#| eval: true
#| fig-show: hide
gc_profile_df <- data.frame(
  Position   = 1:length(hs_gc_profile),
  GC_Content = c(hs_gc_profile, dr_gc_profile),
  Species    = rep(c("Human", "Zebrafish"), each = length(hs_gc_profile))
)

p <- ggplot(gc_profile_df, aes(x = Position, y = GC_Content, color = Species)) +
  geom_line() +
  scale_color_manual(values = c("Human" = "red", "Zebrafish" = "blue")) +
  geom_hline(yintercept = gc_hs_chr1, color = "red",  linetype = "dashed") +
  geom_hline(yintercept = gc_dr_chr1, color = "blue", linetype = "dashed") +
  labs(title = "GC content profile around TSS",
       subtitle = "Dashed lines: chromosome-1 GC baseline per species",
       x = "Position relative to TSS (bp)",
       y = "Mean GC content") +
  theme_minimal() +
  theme(legend.position = "top")
```

## The same figure with ggplot2 — plot

```{r}
#| echo: false
#| include: true
#| fig-width: 11
#| fig-height: 6
print(p)
```

> 📝 **Task 2** — rescale the x-axis so the TSS sits at 0 and the axis runs
> from roughly -1000 to +1000 bp.

## Summary

-   **BSgenome packages** make whole genomes available as lazily-loaded R objects.
-   **`import()`** from rtracklayer reads BED / GFF / BigWig into `GRanges`.
-   Seqname conventions differ — always align them before `getSeq()`.
-   **`getSeq()`** extracts sequences by `GRanges`, respecting strand.
-   **`letterFrequencyInSlidingView()`** computes per-position letter counts along a sequence.
-   Wrapping the pipeline into a function (`get_N_profile()`) makes the same analysis reusable for any letter set, window, and species.

Final exercise: repeat with `letters = "CG"` to build a CpG profile — expect a much sharper peak than GC.
