---
title: "Introduction to Base Graphics in R"
format: revealjs
editor: visual
---
<style>
.reveal {
  font-size: 160%;
}
</style>



### Scatter Plot

```{r}
#| echo: true
#| include: true
# Generate sample data
set.seed(123); x <- rnorm(50)+1:50; y <- rnorm(50)+1:50

# Basic scatter plot
plot(x, y,
     main = "Scatter Plot",
     xlab = "X-axis",
     ylab = "Y-axis",
     pch = 19,       # solid circle
     col = "blue")
```
## Line Plot
```{r}
#| echo: true
#| include: true
x <- 0:10
y <- x^2

plot(x, y,
     type = "l",
     main = "Line Plot",
     xlab = "X-axis",
     ylab = "Y-axis",
     col = "red")
```
## Plot options
```{r}
#| echo: true
#| include: true
par(mfrow = c(2, 2))
plot(x, y, type = "l", main = "type = 'l'")
plot(x, y, type = "p", main = "type = 'p'")
plot(x, y, type = "b", main = "type = 'b'")
plot(x, type = "h", main = "type = 'h'")
```
## Adding more data to the plot
ToothGrowth data set contains the results from an experiment to compare the effect of two supplements on tooth growth in 60 guinea pigs.
 A data frame with 60 observations on 3 variables.

- [,1]  len   numeric  Tooth length
- [,2]  supp  factor   Supplement type (VC or OJ).
- [,3]  dose  numeric  Dose in milligrams/day

```{r}
#| echo: false
#| include: true
library(knitr)
data(ToothGrowth)
kable(head(ToothGrowth))
```
## Adding more data to the plot - simple plot
```{r}
#| echo: true
#| include: true
plot(ToothGrowth$dose, ToothGrowth$len,
     main = "Tooth Growth",
     xlab = "Dose (mg/day)",
     ylab = "Tooth length",
     pch = 19,       # solid circle
     col = "blue")
```
## Adding more data to the plot - by supplement type
```{r}
#| echo: true
#| include: true
tg_vc <- ToothGrowth[ToothGrowth$supp == "VC", ]
tg_oj <- ToothGrowth[ToothGrowth$supp == "OJ", ]
plot(tg_vc$dose, tg_vc$len,
     main = "Tooth Growth by Supplement Type",
     xlab = "Dose (mg/day)",
     ylab = "Tooth length",
     pch = 19,       # solid circle
     col = "blue")
points(tg_oj$dose, tg_oj$len, pch = 19, col = "red")
legend("bottomright", legend = c("VC", "OJ"), col = c("blue", "red"), pch = 19)

```
## `plot` function vs. `points` function
- `plot` function creates a new plot
- `points` function adds points to an existing plot
- `lines` function adds lines to an existing plot
  - `lines` function is like `points(type = "l")`
```{r}
#| echo: false
#| include: true
par(mfrow = c(1,3))
xx <- 1:10
yy <- c(1,3,2,5,4,7,6,9,8,10)
plot(xx,yy, type = "b", main = "single plot")
plot(xx,yy, type = "b", main = "plot + points")
points(xx,yy/2, type = "p", col = "red")
plot(xx,yy, type = "b", main = "plot + points + lines")
points(xx,yy/2, type = "l", col = "red")
lines(xx,yy*2, col = "blue")
```
## Plotting categorical data
Here `len ~ supp` is a **formula** that tells R to plot `len` on the y-axis and `supp` on the x-axis.
```{r}
#| echo: true
#| include: true
boxplot(len ~ supp, data = ToothGrowth,
        main = "Tooth Growth by Supplement Type",
        xlab = "Supplement Type",
        ylab = "Tooth length"
)
```

## Plotting categorical data
Input for `boxplot` function can be a formula, a matrix, data frame, or a list.
```{r}
#| echo: false
#| include: true
example_list <- list(a = rnorm(100), b = rnorm(100,3, 0.5), c = rnorm(100, -2, 5))
boxplot(example_list)

```
## Categorical data - `stripchart` function
```{r}
#| echo: true
#| include: true
stripchart(len ~ supp, data = ToothGrowth,
           main = "Tooth Growth by Supplement Type",
           xlab = "Supplement Type",
           ylab = "Tooth length",
           col = "blue",
           vertical = TRUE
)
```
Stripchart is a one-dimensional scatter plot that represents the distribution of a continuous variable for different levels of a categorical variable. It can be used for similar data like boxplot.

## Categorical data - `stripchart` function
What is points overlap? We can use `jitter` to spread the points out.
```{r}
#| echo: true
#| include: true
stripchart(len ~ supp, data = ToothGrowth,
           main = "Tooth Growth by Supplement Type",
           xlab = "Supplement Type",
           ylab = "Tooth length",
           col = "blue",
           method = "jitter",
           vertical = TRUE
)
```

##  Bar plot vs Box plot
- Use boxplots when you need to display the full distribution, including medians, quartiles, and outliers
- Boxplots are ideal for comparing the spread and distribution across multiple groups
- Bar charts can mask the underlying data distribution; opt for boxplots when the detailed spread and potential outliers are important
- Use Bar charts when you want to show a distribution of data points or perform a comparison of metric values across different subgroups of your data

```{r}
#| echo: false
#| include: true
#| fig.width: 15
par(mfrow = c(1,4))
counts <- table(mtcars$cyl)
boxplot(mpg ~ cyl, data = mtcars,
        main = "Boxplot of MPG by Cylinders",
        xlab = "Number of Cylinders",
        ylab = "Miles per Gallon",
        col = "lightblue")
barplot(counts,
        main = "Bar Chart of Cylinders",
        xlab = "Number of Cylinders",
        ylab = "Frequency",
        col = "lightgreen")

```
## Bar plot example
```{r}
#| echo: true
#| include: true
counts <- table(mtcars$gear) # frequency table
par(mfrow = c(1,2))
barplot(counts,
        main = "Bar Chart of Gears",
        xlab = "Number of Gears",
        ylab = "Frequency",
        col = "lightgreen")
# multiple bar plots
counts <- table(mtcars$am, mtcars$gear)
barplot(counts,
        main = "Bar Chart of Gears by Transmission",
        xlab = "Number of Gears",
        ylab = "Frequency",
        col = c("lightblue", "lightgreen"))
```
## Histogram
Function `hist` takes a vector of values and computes a histogram.

```{r}
#| echo: true
#| include: true
x2 <- rnbinom(100, mu = 10, size = 10)
hist(x2,
     main = "Histogram of Negative Binomial Distribution",
)
```

## Histogram
In default settings, `hist` calculates the number of bins automatically. You can specify the number of bins using the `breaks` argument.
```{r}
#| echo: true
#| include: true
par(mfrow=c(1,2))
hist(x2,
     main = "breaks = 20",
     breaks = 20
)
hist(x2,
     main = "breaks = 1:40",
     breaks = 0:40
)
```
## Customizing plots
```{r}
#| echo: true
#| include: true
par(mfrow = c(2,3))
plot(x,y, main='default plot')
plot(x,y, col="#FF0000", main ="col set point color")
plot(1:20, col=rainbow(20), main ='col is a vector os colors')
plot(1:20, pch=1:20, main = "pch set point shape")
plot(x,y, cex = 3, main = "cex set point size")
plot(x,y, main = "plot with lines and grid")
abline(0, 2) # line with intercept 0 and slope 1
grid()

```
## Customizing plots
```{r}
#| echo: true
#| include: true
par(mfrow = c(2,3))
plot(x,y, main='default plot', type='b')
plot(x,y, xlim=c(0, 30), ylim=c(0, 20), type='b', main = "setting limits of x and y axes")
plot(x,y, log='y', type='b', main = "log scale on y-axis")
plot(x,y, log='x', type='b', main = "log scale on x-axis")
plot(x,y, log='xy', type='b', main = "log scale on both axes")
```

## Adding legends and annotations

```{r}
#| echo: true
#| include: true
plot(x, y, pch = 19, col = "blue")
points(x + 0.5, y + 0.5, col = "red", pch = 17)
# Add a legend to differentiate groups
legend("topleft",
       legend = c("Group 1", "Group 2"),
       col = c("blue", "red"),
       pch = c(19, 17))
# Add a mathematical annotation using expression()
mtext(expression(alpha + beta == gamma), side = 3, cex =2)
```
## Combining multiple plots
Regular grid layout using `par` function
```{r}
#| echo: true
#| include: true
par(mfrow = c(2, 3))  # vector specifying the number of rows and columns
plot(x, y, main = "plot 1", col = "1")
plot(x, y, main = "plot 2", col = "2")
plot(x, y, main = "plot 3", col = "3")
plot(x, y, main = "plot 4", col = "4")
plot(x, y, main = "plot 5", col = "5")
plot(x, y, main = "plot 6", col = "6")
```
## Combining multiple plots
Custom grid layout using `layout` function
```{r}
#| echo: true
#| include: true
layout(matrix(c(1, 2, 4, 3), nrow=2), width = c(2,1), height = c(1,2))
layout.show(4)

```
## Combining multiple plots
Custom grid layout using `layout` function
```{r}
#| echo: true
#| include: true
layout(matrix(c(1, 2, 4, 3), nrow=2), width = c(2,1), height = c(1,2))
hist(x2, main = "1st plot")
plot(x, y, main = "2nd plot")
hist(rnorm(100), main = "3rd plot")
```

## Exporting plots
- `png` for bitmap graphics

```{r}
#| echo: true
#| include: true

png("outputs/plot1.png", width = 800, height = 600) # by defaults units are pixels
# whatever is plotted between `png` call and `dev.off` will be saved in the file
plot(x, y, main = "Scatter Plot", col = "blue")
dev.off()
```
- Alternative output functions for bitmap graphics: `jpeg`, `bmp`, `tiff`
- The output always include single plot. If you want to save multiple plots, use `par` or `layout` functions.

## Exporting plots
- `pdf` for vector graphics - also suitable for multi-page documents
- `svg` for vector graphics
Vector graphics is suitable if you want to do additional editing in a vector graphics editor like Inkscape or Adobe Illustrator.

```{r}
#| echo: true
#| include: true
pdf("outputs/plot2.pdf", width = 8, height = 6)
plot(x, y, main = "Scatter Plot", col = "blue") # first page
hist(x2, main = "Histogram")                    # second page
plot(x, y, main = "Line Plot", col = "red")     # third page
dev.off()
```
## `par` function
- `par` function is used to set large number of graphical parameters
- It can be used to set the layout of multiple plots
   - `par(mfrow = c(2, 2))` will create a 2x2 grid layout
- `par` function can be used to set many graphical parameters like margins, font size, font family, etc.
   - `par(mar = c(5, 4, 4, 2) + 0.1)` will set the margins of the plot
- `par` function can be used to set graphical parameters for text, lines, points, etc.
   - `par(cex = 1.5)` will set the size of text to 1.5 times the default size
- `par` function can be used to set graphical parameters for colors, line types, line widths, etc.
   - `par(lty = 2)` will set the line type to dashed
- see `?par` for more details

## Summary
- Scatter plot: `plot(x, y)`
- Line plot: `plot(x, y, type = "l")`
- Box plot: `boxplot(len ~ supp, data = ToothGrowth)`
- Bar plot: `barplot(counts)`
- Histogram: `hist(x)`
- Stripchart: `stripchart(len ~ supp, data = ToothGrowth)`
- Exporting plots: `png()`, `pdf()`
- Customizing plots: `par()`
- Combining plots: `par(mfrow = c(2, 2))`, `layout()`