---
title: "R introduction"
format: revealjs
editor: visual
---

# Control Structures

- `if` and `else`: testing a condition and acting on it

- `for`: execute a loop a fixed number of times

- `while`: execute a loop _while_ a condition is true

- `repeat`: execute an infinite loop (must `break` out of it to stop)

- `break`: break the execution of a loop

- `next`: skip an interation of a loop




## Control Structures: if-else

```{R}
#| echo: true
#| include: true
#| eval: false
if (<condition>) {
  <statements>
  # do something
} else {
  <statements>
  # do something else
}
```

Example:
```{r}
#| echo: true
#| include: true
x <- 10
if (x > 5) {
  print("x is greater than 5")
} else {
  print("x is less than or equal to 5")
}
```
## Control Structures: for loop
For loops are used to iterate over a sequence of values.
```{R}
#| echo: true
#| include: true
#| eval: false
for (i in <sequence>) {
  <statements>
}
```
Example:
```{r}
#| echo: true
#| include: true
for (i in 1:5) {
  print(i)
}
```
## Control Structures: for loop
For loops can be used to iterate over a vector.

```{r}
#| echo: true
#| include: true
x <- seq(0, 1, by = 0.3)
for (i in x) {
  print(i)
}
```
Alternativelly we can iterate over indices.
```{r}
#| echo: true
#| include: true
for (i in seq_along(x)) {
  cat(i, x[i], "\n")
}
```
## Nested loops
```{r}
#| echo: true
#| include: true
mat <- matrix(1:6, nrow = 3)
print(mat)
for (i in seq_len(nrow(mat))) {
  for (j in seq_len(ncol(mat))) {
    print(mat[i, j])
  }
}
```
## Other types of loops
- `while` loops: execute a loop while a condition is true
- `repeat` loops: execute an infinite loop
- `next` skips the current iteration
- `break` exits the loop

```{r}
#| echo: true
#| include: true
#| eval: false
while (<condition>) {
  <statements>
}

repeat {
  <statements>
  if (<condition>) {
    break
  }
}
```
## Vectorized operations
- R is optimized for vectorized operations
- The most efficient way to perform operations in R is to avoid loops
```{r}
#| echo: true
#| include: true
x <- 1:5
y <- 6:10
z <- x + y
print(z)
```
```{r}
#| echo: true
#| include: true
z <- numeric(length(x))
for (i in seq_along(x)) {
    z[i] <- x[i] + y[i]
}
print(z)
```
## Apply family of functions
- `apply()`, `lapply()`, `sapply()`, `tapply()`, `mapply()`
- These functions are used to apply a function to a vector or list
- They are more efficient than loops
```{r}
#| echo: true
#| include: true
#| eval: false
lapply(<list>, <function>)
```
- *function* is a function that takes a single argument and returns a single value.
- *function* can be a built-in function or a custom function.
- *function* is 'applied' to each element of the list.


## Example: `lapply()`

```{r}
#| echo: true
#| include: true
example_list <- list(a = 1:3, b = 4:6, c = 7:9)
m <- lapply(example_list, mean)
print(m)
```
```{r}
#| echo: true
#| include: true
m <- numeric(length(example_list))
for (i in seq_along(example_list)) {
    m[i] <- mean(example_list[[i]])
}
print(m)
```

## `sapply`
`sapply` is a simplified version of `lapply` that returns a vector or matrix.
```{r}
#| echo: true
#| include: true
m <- sapply(example_list, mean)
print(m)
r <- sapply(example_list, range)
print(r)
```
## `apply`
- `apply` is used to apply a function to the rows or columns of a matrix

```{r}
#| echo: true
#| include: true
mat <- matrix(1:6, nrow = 3)
print(mat)
row_sums <- apply(mat, 1, sum)
print(row_sums)
col_sums <- apply(mat, 2, sum)
print(col_sums)
```
## `mapply`
- `mapply` is a multivariate version of `sapply`
- It applies a function to the first elements of each argument, the second elements, and so on.

```{r}
#| echo: true
#| include: true
x <- 1:3
y <- 4:6
m <- mapply(rep, x, y)
print(m)
```
## `tapply`
- `tapply` is used to apply a function to subsets of a vector

`x` data frame:
```{r}
#| echo: false
#| include: true
library(knitr)
x <- read.csv("../data/sample_data.csv")
kable(head(x))

```
## `tapply`
calulate the mean of Leaf_Count in each Treatment group
```{r}
#| echo: true
#| include: true
tapply(x$Leaf_Count, x$Treatment, mean)
```
For each level of `Treatment`, `tapply` calculates the mean of `Leaf_Count`.

## Functions
- Functions are blocks of code that perform a specific task
- Functions take arguments and return a value
- Functions can be built-in, part of a package, or user-defined

User-defined function:
```{r}
#| echo: true
#| include: true
#| eval: false
my_function <- function(x) {
   <do some stuff with x>
   return(some value)
}
```
## Functions
Single argument function:
```{r}
#| echo: true
#| include: true
get_odd_numbers <- function(x) {
  odd_numbers <- x[x %% 2 == 1]
  return(odd_numbers)
}
odd_numbers <- get_odd_numbers(1:10)
print(odd_numbers)
```
Multiple argument function:
```{r}
#| echo: true
#| include: true
add_numbers <- function(x, y) {
  return(x + y)
}
```
## Functions
Functions can have default arguments
```{r}
#| echo: true
#| include: true
increment_value <- function(x, inc = 1) {
  return(x + inc)
}
incremented_value <- increment_value(5)
print(incremented_value)
incremented_value <- increment_value(5, 2)
print(incremented_value)
```
