---
title: "Data manipulation and cleaning with Tidyverse"
format:
  revealjs:
    self-contained: true
editor: visual
---
<style>
.reveal {
  font-size: 160%;
}
</style>

## Tidyverse
- **Tidyverse** is a collection of R packages designed for data science.
- It provides a consistent set of functions and syntax for data manipulation and visualization.
- Key packages include `dplyr`, `ggplot2`, `tidyr`, and `readr`.
  - `dplyr`: data manipulation
  - `ggplot2`: data visualization
  - `tidyr`: data cleaning
  - `readr`: data import
- use `install.packages("tidyverse")` to install the package
- use `library(tidyverse)` to load the packages

## Tidyverse `reader` package
- The `readr` package provides functions for reading data into R.
- The functions are similar to base R functions like `read.csv()`, but faster and more user-friendly.
- Common functions:
  - `read_csv()`: reads comma-separated files.
  - `read_tsv()`: reads tab-separated files.
  - `read_delim()`: reads files with custom delimiters.
  - `read_fwf()`: reads fixed-width files.
  - `read_table()`: reads tabular data.
- returns a `tibble`, a modern version of a data frame.
## Sample data

```{r}
#| echo: true
#| include: true
library(tidyverse)

# 1. Basic Data Inspection
# Create a small tibble simulating a gene expression dataset.
df <- tibble(
  `Gene Description` = c("Protein kinase C (alpha)", "Ribosomal protein L2", "ATP synthase", "Control gene"),
  `Gene Accession Number` = c("Gene1", "Gene2", "Gene3", "AFFX-Control"),
  Sample1 = c(5.2, 3.4, 7.8, 1.2),
  call1 = c("P", "A", "P", "A"),
  Sample2 = c(5.5, 3.8, 8.0, 1.3),
  call2 = c("P", "A", "P", "A")
)
```

## Sample data (cont'd)
```{r}
#| echo: true
#| include: true
# Inspect dimensions (number of rows and columns)
print(dim(df))  # e.g., 4 rows, 6 columns

# Print column names
print(colnames(df))

# Show the first few rows of the dataset
print(head(df))
```
## Using the Pipe Operator: `%>%`
- The `%>%` operator passes the result of one function into the next.
```{r}
#| echo: true
#| include: true
s <- rnorm(100)
m <- mean(s)

# Using the pipe operator
m <- s %>% mean()
# or
m <- rnorm(100) %>% mean()
# is equivalent to
m <- mean(rnorm(100))

```


## Using the Pipe Operator (%>%)
- The `%>%` operator passes the result of one function into the next.
- Here, we select the "Gene Description" column and then display its head.

```{r}
#| echo: true
#| include: true
df %>%
  select(`Gene Description`) %>%
  head()

# This is equivalent to
head(select(df, `Gene Description`))

```
## `select()` function (dplyr package)

- The `select()` function is used to select columns from a data frame.
- It can be used to reorder columns or to exclude columns.
- Syntax: `select(data, column1, column2, ...)`
- Use `:` to select a range of columns.
- Use `-` to exclude columns.
- Use `everything()` to include all columns.
- Use `starts_with()`, `ends_with()`, `contains()`, `matches()`, `one_of()` to select columns based on name patterns.
- see `?select` for more details.

## `select()` function examples

```{r}
#| echo: true
#| include: true
# Select columns by name
df %>%
  select(`Gene Description`, `Gene Accession Number`) %>%
  head()
# remove columns using `-`
df %>%
  select(-`Gene Description`, -`Gene Accession Number`) %>%
  head()
# remove columns using `-` and contains()
df %>%
  select(-contains("call")) %>%
  head()
```
## Filtering rows with `filter()` function
- `filter()` is used to select rows based on conditions.
- Syntax: `filter(data, condition)`
- Use logical operators like `==`, `!=`, `>`, `<`, `>=`, `<=`, `%in%`, `&`, `|`, `!` to define conditions.
- Use `between()` to filter rows within a range.
```{r}
#| echo: true
#| include: true

kinase_df <- df %>%
  filter(str_detect(`Gene Description`, "kinase"))
print(kinase_df)

```
## Creating new columns with `mutate()` function
- `mutate()` is used to create new columns or modify existing columns.
- Syntax: `mutate(data, new_column = expression)`

```{r}
#| echo: true
#| include: true
df_extended <- df %>%
  mutate(IsControl = str_detect(`Gene Accession Number`, "^AFFX"))
print(df_extended)
```
## Using `mutate()` to create new columns

- Calculate the average expression across Sample1 and Sample2.
- We select only the numeric sample columns.

```{r}
#| echo: true
#| include: true

df_extended <- df_extended %>%
  mutate(AvgExpr = rowMeans(select(., Sample1, Sample2), na.rm = TRUE))
print(df_extended)
```
## Grouping and summarizing with `group_by()` and `summarize()`

- `group_by()` is used to group data by one or more variables.
- Syntax: `group_by(data, column1, column2, ...)`
- `summarize()` is used to calculate summary statistics for each group.
- Syntax: `summarize(data, new_column = function(column))`
- Un-group data using `ungroup()`.

```{r}
#| echo: true
#| include: true
by_cyl <- mtcars %>% group_by(cyl)
# grouping does not change the data frame (it just adds a grouping attribute)
print(by_cyl)
```
## Grouping changes the behavior of functions

- When a data frame is grouped, functions like `summarize()` will calculate summary statistics for each group.
- Here, we calculate the mean `mpg` for each number of cylinders.

```{r}
#| echo: true
#| include: true
by_cyl <- mtcars %>% group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))
print(by_cyl)
by_cyl_gear <- mtcars %>% group_by(cyl, gear) %>%
  summarize(mean_mpg = mean(mpg))
print(by_cyl_gear)
```
## Grouping (cont'd)
```{r}
#| echo: true
#| include: true
control_summary <- df_extended %>%
  group_by(IsControl) %>%
  summarize(mean_expr = mean(AvgExpr), sd_expr = sd(AvgExpr))
print(control_summary)
```

## Reshaping data with `pivot_longer()` and `pivot_wider()` functions

- `pivot_longer()` is used to convert wide data to long data.
- `pivot_wider()` is used to convert long data to wide data.
- Wide data has multiple columns for each variable.
- Long data has multiple rows for each variable.

## Wide format example
```{r}
#| echo: true
#| include: true
wide_df <- tibble(
  ID = c(1, 2, 3,4),
  Gene = c("Gene1", "Gene2", "Gene3", "Gene4"),
  Expr1 = c(5.2, 3.4, 7.8, 1.2),
  Expr2 = c(5.5, 3.8, 8.0, 1.3),
  Expr3 = c(5.5, 3.8, 8.0, 1.3),
  Expr4 = c(5.5, 3.8, 8.0, 1.3),
  Expr5 = c(5.5, 3.8, 8.0, 1.3)
)
print(wide_df)
```
## Long format example
- Function `pivot_longer()` is used to convert wide data to long data.
- syntax: `pivot_longer(data, cols, names_to, values_to)`
- `cols` is a selection of columns to convert to long format.
- `names_to` is the name of the new column that will store the column names.
- `values_to` is the name of the new column that will store the values.
```{r}
#| echo: true
#| include: true
long_df <- wide_df %>%
  pivot_longer(cols = starts_with("Expr"),
               names_to = "Sample",
               values_to = "Expression")
print(long_df)
```
## Converting long data to wide data
- Function `pivot_wider()` is used to convert long data to wide data.
- syntax: `pivot_wider(data, names_from, values_from)`
- `names_from` is the name of the column that will be used to create new columns.
- `values_from` is the name of the column that will be used to fill the new columns.
```{r}
#| echo: true
#| include: true
wide_df <- long_df %>%
  pivot_wider(names_from = "Sample",
              values_from = "Expression")
print(wide_df)
```
## String manipulation with functions from `stringr` package
- The `stringr` package provides functions for string manipulation.
- The most common functions:
  - `str_detect()`: detect patterns in strings.
  - `str_replace()`: replace patterns in strings.
  - `str_split()`: split strings based on patterns.
  - `str_sub()`: extract substrings.
  - `str_trim()`: remove leading and trailing whitespace.
  - `str_to_lower()`, `str_to_upper()`: convert strings to lower or upper case.
  - `str_c()`: concatenate strings.
  - `str_remove()`, `str_remove_all()`: remove patterns from strings.

## `stringr` package examples
```{r}
#| echo: true
#| include: true
library(stringr)
# detect patterns in strings
str_detect("Hello, world!", "world")
# replace patterns in strings
str_replace("Hello, world!", "world", "universe")
# split strings based on patterns
str_split("Hello, world!", ",")
# extract substrings
str_sub("Hello, world!", 3, 7)
# remove leading and trailing whitespace
str_trim("  Hello, world!  ")
# remove patterns from strings
str_remove("Hello, world!", "world")
```
## Regular expressions : defining search patterns
- Basic regex syntax:
  - `.`: matches any character.
  - `^`: matches the start of a string.
  - `$`: matches the end of a string.
  - `[]`: matches any character inside the brackets.
  - `|`: matches either the expression before or after the pipe.
  - `*`: matches zero or more occurrences of the preceding character.
  - `+`: matches one or more occurrences of the preceding character.
  - `?`: matches zero or one occurrence of the preceding character.
  - `{n}`: matches exactly n occurrences of the preceding character.
  - `{n,}`: matches n or more occurrences of the preceding character.
  - `{n,m}`: matches between n and m occurrences of the preceding character.
  - `\\`: escape character.
  - `\\d`: matches any digit.
  - `\\w`: matches any word character.

## Regular expressions examples
```{r}
#| echo: true
#| include: true
# 1. Dot (.)
# Matches any character except a newline.
str_detect("cat", "c.t")       # TRUE: 'c' + any character ('a') + 't'
str_detect("ct", "c.t")        # FALSE: missing a character between 'c' and 't'

# 2. Start (^) and End ($)
# ^ matches the start of a string, $ matches the end.
str_detect("Hello world", "^Hello")  # TRUE: string starts with "Hello"
str_detect("Hello world", "world$")   # TRUE: string ends with "world"

# 3. Character Class ([])
# Matches any single character listed inside the brackets.
str_extract("hello", "[aeiou]")         # "e": extracts the first vowel found
str_extract_all("hello", "[aeiou]")
```
## Regular expressions examples (cont'd)
```{r}
#| echo: true
#| include: true
# 4. Alternation (|)
# Matches either the expression before or after the pipe.
str_detect("I like cats", "cat|dog")      # TRUE: "cat" is found
str_detect("I like dogs", "cat|dog")      # TRUE: "dog" is found
str_detect("I like birds", "cat|dog")     # FALSE: neither "cat" nor "dog" is found

# 5. Asterisk (*)
# Matches zero or more occurrences of the preceding element.
str_detect("ac", "ab*c")    # TRUE: zero 'b's between 'a' and 'c'
str_detect("abbc", "ab*c")  # TRUE: one or more 'b's between 'a' and 'c'
# 6. Plus (+)
# Matches one or more occurrences of the preceding element.
str_detect("abc", "ab+c")   # TRUE: one 'b' present
str_detect("abbc", "ab+c")  # TRUE: two 'b's present
str_detect("ac", "ab+c")    # FALSE: requires at least one 'b'
```
## Regular expressions examples (cont'd)
```{r}
#| echo: true
#| include: true

# 7. Question Mark (?)
# Matches zero or one occurrence of the preceding element.
# Example: "colou?r" matches both "color" and "colour"
str_detect("color", "colou?r")    # TRUE
str_detect("colour", "colou?r")   # TRUE

# 8. Exact Count ({n})
# Matches exactly n occurrences of the preceding element.
str_detect("aaa", "a{3}")   # TRUE: exactly three 'a's
str_detect("aa", "a{3}")    # FALSE: only two 'a's

# 9. Minimum Count ({n,})
# Matches n or more occurrences.
str_detect("aa", "a{2,}")    # TRUE: at least two 'a's present
str_detect("a", "a{2,}")     # FALSE: only one 'a'
```
## Regular expressions examples (cont'd)
```{r}
#| echo: true
#| include: true
# 10. Range Count ({n,m})
# Matches between n and m occurrences (inclusive).
str_detect("aa", "a{2,4}")    # TRUE: two 'a's is within the range
str_detect("aaaa", "a{2,4}")  # TRUE: four 'a's is within the range
str_detect("aaaaa", "a{2,4}") # FALSE: five 'a's exceeds the range

# 11. Escape Character (\\)
# Use double backslashes to escape special characters.
# For example, to match a literal period (.) you must escape it.
str_detect("a.b", "a\\.b")   # TRUE: matches literal "a.b"
str_detect("aXb", "a\\.b")   # FALSE: "X" is not a literal period
```
## Regular expressions examples (cont'd)
```{r}
#| echo: true
#| include: true
# 12. Digit (\\d)
# Matches any digit (0-9).
str_extract("Order number: 12345", "\\d+")  # "12345": extracts one or more digits

# 13. Word Character (\\w)
# Matches any word character (letters, digits, underscore).
str_extract_all("Hello, world!", "\\w+")   # c("Hello", "world"): extracts words
```
## Testing regular expressions

- Use online tools like [regex101](https://regex101.com/) to test regular expressions.

