---
title: "Using phylogenetic trees in R"
format:
  revealjs:
    self-contained: true
    slide-number: true
editor: visual
---

<style>
.reveal {
  font-size: 160%;
}
</style>

```{r}
#| echo: false
#| include: false
library(ape)
library(phangorn)
library(treeio)
library(ggtree)
library(dplyr)
library(tidyr)
library(tibble)
```

# Phylogenetic Analysis in R with `ape` and `treeio`

Phylogenetic trees are key to understanding evolutionary relationships. In R, the `ape` and `treeio` packages provide tools to import, manipulate, and visualize phylogenetic trees.

These packages can be used to:

- Load trees from Newick/Nexus formats
- Create and manipulate trees (root, prune, extract clades)
- Visualize trees with base ape (`plot.phylo`) and `ggtree`
- Annotate trees with metadata and heatmaps

---

## Packages installation

```{r}
#| eval: false
#| echo: true
#| include: true
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if (!requireNamespace("ape", quietly = TRUE))
  install.packages("ape", update = FALSE)
if (!requireNamespace("treeio", quietly = TRUE))
  BiocManager::install("treeio", update = FALSE)
if (!requireNamespace("ggtree", quietly = TRUE))
  BiocManager::install("ggtree", update = FALSE)
if (!requireNamespace("phangorn", quietly = TRUE))
  install.packages("phangorn", update = FALSE)
```

## Load packages

```{r}
#| echo: true
#| include: true
library(ape)
library(phangorn)
library(treeio)
library(ggtree)
library(dplyr)
library(tidyr)
library(tibble)
```

# Section 1: ape basics

## `ape` package

`ape` (Analysis of Phylogenetics and Evolution) stores trees as `phylo` objects and provides:

- `read.tree()` / `write.tree()` — Newick I/O
- `rtree()` — random tree generation
- `root()`, `drop.tip()`, `ladderize()` — tree manipulation
- `plot.phylo()` — base R tree visualization

---

## Functions to read and write trees

`read.tree()` reads Newick format:
`((A:0.1,B:0.2):0.3,C:0.4);`

```{r}
#| echo: true
#| include: true
newick_tree <- "((A:0.1,B:0.2):0.3,C:0.4);"
tree <- read.tree(text = newick_tree)
plot(tree)
axis(1)
```

## Plotting trees with `ape` {.smaller}

`plot()` on a `phylo` object calls `plot.phylo()`.

```{r}
#| echo: true
#| include: true
#| fig-height: 3
tr <- read.tree(text = "((Pan:5,Human:5):2,Gorilla:7);")
par(mfrow = c(2, 3))
plot(tr, type = "cladogram", main = "cladogram")
plot(tr, type = "unrooted",  main = "unrooted")
plot(tr, type = "fan",       main = "fan")
plot(tr, type = "radial",    main = "radial")
plot(tr, type = "tidy",      main = "tidy")
```

## Loading tree from file

```{r}
#| echo: true
#| include: true
tr_dentist <- read.tree("../data/phylogenetics/HIV_dentist.nhx")
plot(tr_dentist)
```

## Properties of `phylo` class

```{r}
#| echo: true
#| include: true
tr <- rtree(5)   # random tree with 5 tips
print(tr)
class(tr)
names(tr)
```

## Properties of `phylo` class — `edge`

- `edge`: matrix with two columns; each row is one branch (parent → child)

```{r}
#| echo: true
#| include: true
plot.phylo(tr, type = "tidy", show.tip.label = TRUE, cex = 1)
print(tr$edge)
```

## Properties of `phylo` class — `tip.label` and `edge.length`

```{r}
#| echo: true
#| include: true
print(tr$tip.label)
print(tr$edge.length)
plot(tr, type = "tidy", show.tip.label = TRUE, cex = 1)
axis(1)
```

## Properties of `phylo` class — numbered diagram {.smaller}

```{r}
#| echo: true
#| include: true
tr$tip.label  <- as.character(1:Ntip(tr))
tr$node.label <- as.character((Ntip(tr) + 1):(Ntip(tr) + Nnode(tr)))
cat("Edge matrix:\n")
print(tr$edge)
plot(tr, type = "phylogram", use.edge.length = FALSE,
     show.tip.label = TRUE, show.node.label = TRUE,
     label.offset = 0.05, cex = 1, edge.color = "darkgray",
     main = "Tree with Numbered Tips, Nodes, and Edges")
edgelabels(text = seq_len(nrow(tr$edge)),
           frame = "none", adj = 0.5, cex = 1, col = "red")
```

# Section 2: Tree manipulation

## Manipulating trees

- `is.rooted()` — check if rooted
- `unroot()` — remove the root
- `root()` — re-root at a tip or node

```{r}
#| echo: true
#| include: true
tips <- c("Outgroup", paste0("Species", 1:5))
set.seed(123)
unrooted_tree <- rtree(6, tip.label = tips, rooted = FALSE)
is.rooted(unrooted_tree)
plot(unrooted_tree, main = "Unrooted tree")
```

## Rooting a tree

```{r}
#| echo: true
#| include: true
rooted_tree <- root(unrooted_tree, outgroup = "Outgroup", resolve.root = TRUE)
is.rooted(rooted_tree)
plot(rooted_tree, main = "Rooted tree")
```

## Rooting a tree

```{r}
#| echo: true
#| include: true
rooted_tree <- root(unrooted_tree, outgroup = "Species1", resolve.root = TRUE)
plot(rooted_tree, main = "Rooted tree at Species1")
```

## Ladderizing a tree

- Reorders branches so the tree has a tidy ladder-like appearance.
- Evolutionary relationships are unchanged; only the visual order of tips changes.

```{r}
#| echo: true
#| include: true
tree2 <- rtree(8)
par(mfrow = c(1, 2))
plot(tree2,            main = "Original ordering")
plot(ladderize(tree2), main = "After ladderize()")
```

## Pruning and dropping tips

```{r}
#| echo: true
#| include: true
tr <- rtree(10)
par(mfrow = c(1, 2))
plot(tr, main = "Original tree")
pruned_tree <- drop.tip(tr, c("t1", "t2"))
plot(pruned_tree, main = "Pruned tree")
```

# Section 3: Clade extraction

## Extracting clades

- `getMRCA(tree, tips)` — most recent common ancestor node for a set of tips
- `extract.clade(tree, node)` — subtree descended from a node
- `nodepath(tree, from, to)` — sequence of nodes between two nodes

## Understanding node numbering

- Tips: numbered **1 to N**
- Internal nodes: numbered **(N+1) to (N + Nnode)**

```{r}
#| echo: true
#| include: true
set.seed(123)
tr <- rtree(3)
tr$tip.label  <- as.character(1:Ntip(tr))
tr$node.label <- as.character((Ntip(tr) + 1):(Ntip(tr) + Nnode(tr)))
plot(tr, type = "phylogram", use.edge.length = FALSE,
     show.tip.label = TRUE, show.node.label = TRUE,
     label.offset = 0.05, cex = 1, edge.color = "darkgray",
     main = "Numbered Tips, Nodes, and Edges")
edgelabels(text = seq_len(nrow(tr$edge)),
           frame = "none", adj = 0.5, cex = 1, col = "red")
```

## Getting MRCA

```{r}
#| echo: true
#| include: true
mrca_node <- getMRCA(tr, c("1", "2"))
print(mrca_node)
plot(tr, show.tip.label = TRUE, show.node.label = TRUE,
     type = "phylogram", use.edge.length = FALSE,
     tip.color = c(2, 2, 1))
```

## Extracting clades descended from MRCA

```{r}
#| echo: true
#| include: true
clade <- extract.clade(tr, mrca_node)
plot(clade, show.tip.label = TRUE, show.node.label = TRUE,
     type = "phylogram", use.edge.length = FALSE)
```

## Extracting clades descended from MRCA

```{r}
#| echo: true
#| include: true
tr_dentist_rooted <- root(
  tr_dentist,
  outgroup     = which(tr_dentist$tip.label %in% "isolate_ELI/19-301"),
  resolve.root = TRUE
)
MRCA_node        <- getMRCA(tr_dentist_rooted, grep("Dentist", tr_dentist_rooted$tip.label))
tr_dentist_clade <- extract.clade(tr_dentist_rooted, MRCA_node)
par(mfrow = c(1, 2))
color <- grepl("Dentist", tr_dentist_rooted$tip.label) + 1
plot(tr_dentist_rooted, type = "phylogram", use.edge.length = FALSE, tip.color = color)
plot(tr_dentist_clade,  type = "phylogram", use.edge.length = FALSE)
```

# Section 4: Base ape plotting

## Basic plotting with `plot.phylo`

The simplest call: `plot(my_tree)`

Key arguments:
- `type`: `"phylogram"`, `"cladogram"`, `"fan"`, `"unrooted"`, `"radial"`
- `edge.color`, `edge.width`, `tip.color`, `cex`
- `use.edge.length`: `FALSE` draws as cladogram regardless of type

```{r}
#| echo: true
#| include: true
data(bird.orders)
plot(bird.orders, cex = 0.6, main = "Bird Orders (phylogram)")
```

## Basic plotting with `plot.phylo`

```{r}
#| echo: true
#| include: true
plot(bird.orders, type = "fan", cex = 0.7, tip.color = "blue",
     main = "Bird Orders (fan layout)")
```

## Basic plotting with `plot.phylo`

```{r}
#| echo: true
#| include: true
plot(bird.orders, type = "unrooted", cex = 0.7, tip.color = "blue",
     main = "Bird Orders (unrooted layout)")
```

## Adding annotations in base plots

- `tiplabels()` — symbols or text next to tips
- `nodelabels()` — labels at internal nodes (e.g. bootstrap values)
- `edgelabels()` — text on branches

```{r}
#| echo: true
#| include: true
plot(bird.orders, cex = 0.6, main = "Bird Orders (phylogram)")
nodelabels(frame = "circle", cex = 0.5, col = "red")
```

## Setting colors for tip labels

```{r}
#| echo: true
#| include: true
tip_colors <- ifelse(
  bird.orders$tip.label %in% c("Anseriformes", "Galliformes"),
  "red", "black"
)
plot(bird.orders, cex = 1, main = "Bird Orders (phylogram)", tip.color = tip_colors)
```

## Setting colors for tip labels

Set color for clade shared by `Anseriformes` and `Struthioniformes`:

```{r}
#| echo: true
#| include: true
mrca_node        <- getMRCA(bird.orders, c("Anseriformes", "Struthioniformes"))
clade            <- extract.clade(bird.orders, mrca_node)
tip_labels_clade <- clade$tip.label
tip_colors       <- ifelse(bird.orders$tip.label %in% tip_labels_clade, "red", "black")
plot(bird.orders, cex = 1, main = "Bird Orders (phylogram)", tip.color = tip_colors)
```

> 📝 **Task 1** — Plot bird.orders as `"fan"`. Color the MRCA clade of `"Passeriformes"` and `"Piciformes"` blue, all others black.

# Section 5: ggtree

## Visualizing trees with `ggtree`

`ggtree` extends ggplot2 for phylogenetic trees — every ggplot2 layer and theme applies.

```{r}
#| echo: true
#| include: true
nwk  <- system.file("extdata", "sample.nwk", package = "treeio")
tree <- read.tree(nwk)
ggplot(tree, aes(x, y)) + geom_tree() + theme_tree()
```

## Visualizing trees with `ggtree`

`ggtree()` is the shortcut for the above:

```{r}
#| echo: true
#| include: true
ggtree(tree, color = "firebrick", size = 2, linetype = "dotted")
```

## ggtree layouts

```{r}
#| echo: false
#| include: true
#| fig-height: 1.3
#| fig-width: 1.3
#| layout-ncol: 3
set.seed(2017 - 02 - 16)
tree50 <- rtree(50)
ggtree(tree50)
ggtree(tree50, layout = "roundrect")
ggtree(tree50, layout = "slanted")
ggtree(tree50, layout = "ellipse")
ggtree(tree50, layout = "circular")
ggtree(tree50, layout = "fan", open.angle = 120)
ggtree(tree50, layout = "equal_angle")
ggtree(tree50, layout = "daylight")
ggtree(tree50, branch.length = "none")
ggtree(tree50, layout = "ellipse", branch.length = "none")
ggtree(tree50, branch.length = "none", layout = "circular")
ggtree(tree50, layout = "daylight", branch.length = "none")
```

## Phylogenetic Tree Annotation

- `geom_tiplab()` — tip labels
- `geom_nodelab()` — internal node labels
- `geom_treescale()` — scale bar

```{r}
#| echo: true
#| include: true
ggtree(tree) +
  geom_tiplab(size = 3, color = "blue") +
  geom_treescale(x = 0.5, y = 0.5, width = 0.1)
```

## Customizing tree appearance

Since `ggtree` is ggplot2-based:

- `theme_tree()` — minimal tree theme
- `xlim()` / `ylim()` — adjust space around the tree
- `geom_nodepoint()` / `geom_tippoint()` — symbols at nodes/tips
- `geom_text2()` — text at nodes

## Highlighting or annotating clades

- `geom_hilight(node, fill)` — shaded rectangle under a clade
- `geom_cladelabel(node, label)` — text label on a clade

```{r}
#| echo: true
#| include: true
ggtree(bird.orders) +
  geom_tiplab() +
  geom_treescale() +
  geom_hilight(node = 30, fill = "goldenrod", alpha = 0.3) +
  ggplot2::xlim(0, 50)
```

# Section 6: Annotating trees with metadata

## Annotation tree with metadata

Common metadata types:

- **Traits** (body size, habitat)
- **Taxonomic info** (family, order)
- **Geography or host** (pathogen phylogenies)
- **Temporal data** (sample collection year)
- **Statistical results** (dN/dS ratios on branches)

## Combining tree with tip data

```{r}
#| echo: true
#| include: true
tree_example <- rtree(5, tip.label = LETTERS[1:5])
trait_data <- data.frame(
  label  = LETTERS[1:5],
  Status = c("Endangered", "Not Endangered", "Not Endangered",
             "Endangered", "Endangered")
)
trait_data
```

## Attaching data to the tree

`%<+%` attaches a data frame by matching the `label` column:

```{r}
#| echo: true
#| include: true
p <- ggtree(tree_example) %<+% trait_data +
  geom_tiplab(aes(color = Status), size = 4) +
  theme_tree() +
  scale_color_manual(values = c("red", "green"))
p
```

## Attaching data to the tree

```{r}
#| echo: true
#| include: true
p + geom_tippoint(aes(color = Status), size = 3)
```

> 📝 **Task 2** — Assign bird.orders tips to "Waterbirds" (Anseriformes, Gaviiformes, Podicipediformes, Pelecaniformes, Ciconiiformes, Sphenisciformes, Procellariiformes) and "Other". Plot with `ggtree` + `%<+%`, coloring tip points by group.

## Attaching data to the tree — continuous trait

```{r}
#| echo: true
#| include: true
tree_example2 <- rtree(20, tip.label = LETTERS[1:20])
trait_data2 <- data.frame(
  label      = LETTERS[1:20],
  TraitValue = rnorm(20, mean = 5, sd = 2)
)
p2 <- ggtree(tree_example2) %<+% trait_data2 +
  geom_tippoint(aes(color = TraitValue), size = 6)
p2
```

# Section 7: Time-calibrated trees

## Joining tree with metadata using `as.treedata`

When metadata has many columns, use `full_join()` + `as.treedata()`:

```{r}
#| echo: true
#| include: true
#| eval: false
meta <- read_tsv("data/ncov_metadata.tsv")

td <- full_join(as_tibble(tree), meta, by = c("label" = "strain")) |>
        as.treedata()

ggtree(td, mrsd = max(meta$decimal_date), as.Date = TRUE) +
  geom_tippoint(aes(color = Nextstrain_clade))
```

- `as_tibble(tree)` converts the `phylo` object to a tibble with a `label` column
- `full_join()` attaches all metadata columns by tip name
- `as.treedata()` converts back so `ggtree` can access those columns in `aes()`

## Time-calibrated trees with `mrsd`

- `mrsd` = most-recent sampling date (character `"YYYY-MM-DD"` or numeric)
- `as.Date = TRUE` — display x-axis as calendar dates
- `theme_tree2()` — shows the x-axis (time axis)

```{r}
#| echo: true
#| include: true
beast_file <- system.file("examples/MCC_FluA_H3.tree", package = "ggtree")
beast_tree <- read.beast(beast_file)

ggtree(beast_tree, mrsd = "2013-01-01") +
  geom_tiplab(size = 2, align = TRUE, linesize = 0.5) +
  theme_tree2()
```

## Checking branch-length units

`mrsd` only works when branches are in **years** (time-calibrated tree).

```{r}
#| echo: true
#| include: true
#| eval: false
# For a SARS-CoV-2 tree spanning ~6 years this should return ~6
max(node.depth.edgelength(tree))
```

- Result ≈ 6 → branches in years, `mrsd` will work correctly
- Result ≈ 0.001 → branches in substitutions/site → need to time-calibrate first

# Section 8: Heatmaps with `gheatmap`

## Attaching heatmap to the tree

`gheatmap()` glues a **wide matrix** of data cells to the right of a tree:
- Each row = one tip (aligned by **row name**, not row order)
- Each column = one variable (e.g. genotype, mutation presence/absence)

```{r}
#| echo: true
#| include: true
beast_file    <- system.file("examples/MCC_FluA_H3.tree", package = "ggtree")
beast_tree    <- read.beast(beast_file)
genotype_file <- system.file("examples/Genotype.txt", package = "ggtree")
genotype      <- read.table(genotype_file, sep = "\t", stringsAsFactors = FALSE)

p3 <- ggtree(beast_tree, mrsd = "2013-01-01") +
  geom_tiplab(size = 2, align = TRUE, linesize = 0.5) +
  theme_tree2()
head(genotype)
```

## Attaching heatmap to the tree

```{r}
#| echo: true
#| include: true
gheatmap(p3, genotype, offset = 8, width = 0.6,
         colnames = FALSE, legend_title = "genotype") +
  scale_x_ggtree()
```

## `pivot_wider` for `gheatmap`

`gheatmap` requires **wide** format: rows = tips, columns = mutations.

Long → wide with `pivot_wider`:

```{r}
#| echo: true
#| include: true
#| eval: false
key_muts <- c("S:D614G", "S:N501Y", "S:E484K", "S:L452R", "S:P681H")

mut_wide <- spike_long |>
  filter(mutation %in% key_muts) |>
  mutate(present = "present") |>
  pivot_wider(names_from  = mutation,
              values_from = present,
              values_fill = "absent") |>
  column_to_rownames("strain")
```

## `gheatmap` — alignment pitfalls {.smaller}

**Critical:** `gheatmap` aligns by **row name**, not row order.

Two preparations are always needed:

**1. Force every expected column to exist**

```{r}
#| echo: true
#| include: true
#| eval: false
# pivot_wider silently drops columns absent from all tips
for (m in setdiff(key_muts, colnames(mut_wide))) mut_wide[[m]] <- "absent"
mut_wide <- mut_wide[, key_muts]   # enforce column order
```

**2. Force every tree tip to be a row**

```{r}
#| echo: true
#| include: true
#| eval: false
# tips with no key mutations are missing from the pivoted matrix
missing_tips <- setdiff(tree$tip.label, rownames(mut_wide))
if (length(missing_tips) > 0) {
  pad <- as.data.frame(
    matrix("absent", nrow = length(missing_tips), ncol = ncol(mut_wide),
           dimnames = list(missing_tips, colnames(mut_wide)))
  )
  mut_wide <- rbind(mut_wide, pad)
}
```

> 📝 **Task 3** — Inspect `genotype` (loaded above) with `head()` and `class()`. Call `gheatmap()` with `offset = 5`, `width = 0.4`, `colnames = TRUE`. Customize fill colors with `scale_fill_manual()`.
